---
title: "ACS and HAWK Benchmarking"
author: "SL" 
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(openxlsx)
library(tidyverse)
library(gridExtra)
library(scales)
library(gtsummary)
library(knitr)
library(kableExtra)
options(scipen = 999)

```
 
##  1.2. Data read-in 
```{r, echo=F}
dat = read.csv('O:/Korean Health/Data/Codebook/Recoding/Data/DATA_ALL (5.20.2025).csv')

hawk<-dat%>%
  mutate(Age=AGE,
         AgeGe40=ifelse(is.na(AGE),NA,ifelse(AGE>=40,1,0)),
         Male=B1_MALE,
         Married=B2_MARRIED,
         EducGeCollege=B4_MTCOL,
         CurWork=CUR_WORK,
         USBorn=ifelse(is.na(G1),NA,ifelse(G1==1,1,0)),
         USCitizen=ifelse(USBorn%in%c(1)|G5%in%c(1),1,0),
         Intv_Lang=ifelse(LANG_INTERVIEW==1,"KR","EN"),
         Region=ifelse(REGION=="",NA,REGION),
         wave=WAVE,
         num_coupon=COUPON_ISSUED,
  num_recruit=COUPON_USED,
contact_mode=ifelse(is.na(MODE),NA,ifelse(MODE==1,"Email","SMS")),
         any_recruit=ifelse(is.na(COUPON_USED),NA,ifelse(
           COUPON_USED>0,1,0)),
         coupon_type=ifelse(is.na(MODE)|is.na(EXP_COUPON),NA,ifelse(
           MODE==1&EXP_COUPON==1,"Email-Single",ifelse(
             MODE==1&EXP_COUPON==2,"Email-Multiple",ifelse(
               MODE==2&EXP_COUPON==1,"SMS-Verbal","SMS-Image")))),
        num_recruit=COUPON_USED,
         pct_recruit=ifelse(is.na(num_coupon),NA,num_recruit/num_coupon))
```

# 1. Respondent characteristics

```{r, echo=FALSE}
#https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html 

hawk%>%
  select(Age, AgeGe40, Male, Married, 
         EducGeCollege, CurWork, 
         USBorn, USCitizen, Intv_Lang,Region)%>%
  #gtsummary::tbl_summary(missing="no")%>%
  tbl_summary(missing="no")%>%
  #gtsummary::bold_labels() %>%
  bold_labels() %>%
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    linesep = ""
  ) %>%
  kable_styling(
    position = "left",
    latex_options = c("striped", "repeat_header"),
    stripe_color = "gray!15"
  )
```
 
 

# 2. Peer Recruitment
```{r, echo=FALSE} 

hawk%>%
  select(num_coupon, num_recruit)%>%
  tbl_summary(missing="no",
              by=num_coupon,
              label = num_recruit ~ "Recruit_Num")%>%
  modify_header(label ~ "")%>%
  modify_spanning_header(c("stat_1", "stat_2", "stat_3", "stat_4") ~ "**Coupon_Num**") %>%
  bold_labels() %>% 
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    #longtable = TRUE,
    linesep = ""
  ) %>%
  kable_styling(
    position = "left",
    latex_options = c("striped", "repeat_header"),
    stripe_color = "gray!15"
  )
```

+ Coupon use rate:  `r paste0(round(sum(hawk$num_recruit[hawk$num_coupon>0],na.rm=T)/sum(hawk$num_coupon[hawk$num_coupon>0],na.rm=T)*100,1),"%")`

+ Chain length of `r length(unique(hawk$SEED_ID))` seeds

```{r, echo=FALSE}

ChainLength<-hawk%>%
  group_by(SEED_ID)%>%
  summarise(ChainLength=max(wave)+1,
            ChainSampleSize=n())

ChainLength%>%  
  ggplot(aes(x=ChainLength))+
  geom_histogram(color="blue", fill="royalblue1")+
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 20))
```

```{r, echo=FALSE}
ChainLength%>%  
  ggplot(aes(x=ChainSampleSize))+
  geom_histogram(color="red4", fill="tomato1")+
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 20))
```

```{r, echo=FALSE}
ChainLength%>%
  ggplot(aes(x=ChainLength,y=ChainSampleSize))+
  geom_point(position="jitter")+
  geom_smooth()+
  theme(axis.text = element_text(size = 15),
        axis.title = element_text(size = 20))

```

```{r} 
summary(ChainLength$ChainLength)
summary(ChainLength$ChainSampleSize)
```

# 3. Coupon Delivery Mode

## 3.1 Among everyone 

```{r, echo=FALSE}
dat<-hawk%>%
  filter(contact_mode%in%c("Email","SMS")&!is.na(any_recruit))

dat%>%
  select(any_recruit, pct_recruit, num_recruit, contact_mode)%>%
  tbl_summary(missing="no",
              by=contact_mode,
              type = list(c(any_recruit,pct_recruit,num_recruit) ~ 'continuous'),
              statistic = list(all_continuous() ~ "{mean} ({sd})"),
              digits = all_continuous() ~ 2)%>%
    bold_labels()%>%
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    #longtable = TRUE,
    linesep = ""
  ) %>%
  kable_styling(
    position = "left",
    latex_options = c("striped", "repeat_header"),
    stripe_color = "gray!15"
  )

```

- Wilcoxon test
```{r}
wilcox.test(any_recruit~contact_mode,dat)
wilcox.test(pct_recruit~contact_mode,dat)
wilcox.test(num_recruit~contact_mode,dat)
```


## 3.2 Among those for whom mode was randomized
+ There were `r length(hawk$RESPONDENT_ID[hawk$CONTACT_NUM1==2])` respondents who provide both email and phone number.

```{r, echo=FALSE}
dat2<-hawk%>%
  mutate(contact_num1=CONTACT_NUM1)%>%
  filter(contact_mode%in%c("Email","SMS")&!is.na(any_recruit)&contact_num1==2)
```

```{r, results='asis'}
tab_se <- dat2 %>%
  select(any_recruit, pct_recruit, num_recruit, contact_mode) %>%
  pivot_longer(
    cols = c(any_recruit, pct_recruit, num_recruit),
    names_to = "variable",
    values_to = "value"
  ) %>%
  group_by(variable, contact_mode) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    se = sd(value, na.rm = TRUE) / sqrt(sum(!is.na(value))),
    mean_se = sprintf("%.2f (%.2f)", mean, se),
    .groups = "drop"
  ) %>%
  select(variable, contact_mode, mean_se) %>%
  pivot_wider(
    names_from = contact_mode,
    values_from = mean_se
  )
tab_se
```

- Wilcoxon Test 
```{r}
wilcox.test(num_recruit~contact_mode,dat2)
wilcox.test(any_recruit~contact_mode,dat2)
wilcox.test(pct_recruit~contact_mode,dat2)

table(dat2$contact_mode,exclude=NULL)
```
```{r, echo=FALSE} 
#summary(glm(any_recruit~contact_mode,dat2,family="binomial"))
#summary(glm(any_recruit~contact_mode*AgeGe40,dat2,family="binomial"))

#summary(glm(num_recruit~contact_mode+num_coupon,dat2,family="quasipoisson"))
#summary(glm(num_recruit~contact_mode*AgeGe40+num_coupon,dat2,family="quasipoisson"))

```

## 3.3. Mode Interaction with Age
```{r, echo=FALSE, eval=FALSE}


dat2<-dat2%>%
  mutate(Age=ifelse(is.na(AGE),NA, ifelse(AGE<35,"Age<35","Age>=35")))

dat2%>% 
  group_by(contact_mode,Age)%>%
  summarise(m=mean(any_recruit),
            se = sd(any_recruit, na.rm = TRUE) / sqrt(sum(!is.na(any_recruit))))

summary(glm(any_recruit~contact_mode*Age,data=dat2,
            family = binomial(link = "logit")))

```

```{r}
plot_dat <- dat2 %>% 
  group_by(contact_mode, Age) %>%
  summarise(
    m = mean(any_recruit, na.rm = TRUE),
    se = sd(any_recruit, na.rm = TRUE) / sqrt(sum(!is.na(any_recruit))),
    .groups = "drop"
  )

```

```{r}

p_age<-ggplot(plot_dat, aes(x = Age, y = m, 
                            color = contact_mode, 
                            group = contact_mode)) +
  geom_point(
    position = position_dodge(width = 0.4),
    size = 4
  ) +
  geom_errorbar(
    aes(ymin = m - 1.96*se, ymax = m + 1.96*se),
    position = position_dodge(width = 0.4),
    width = 0.2
  )  +
   scale_color_manual(
    values = c(
      "#0072B2",  # blue
      "#D55E00" # vermillion
    )
  )+ 
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1),
    limits = c(0, NA)
  ) +
  labs(
    x = "Recruiter age",
    y = "Any recruits",
    color = "Contact mode"
  ) +
  theme_minimal()+
  theme(
  legend.position = c(0.85, 0.20),
  legend.background = element_rect(fill = "white", color = "gray70"),
      axis.title.x = element_text(size = 16, face = "bold"),
    axis.title.y = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 14)
)

ggsave(filename = "O:/Korean Health/Analysis/SL/NCSES/Figure_AgeContact_NCSES.jpg",
       plot = p_age, width = 8, height = 5, dpi = 100, bg = "white")
```


```{r, echo=FALSE, eval=FALSE}


## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=T,
                      conf.interval=.95) {
datac <- data%>%
      select(measurevar, groupvars)%>%
      group_by(groupvars)%>%
      summarise(N=n(),
                mean=mean (measurevar, na.rm=na.rm),
                sd=sd(measurevar, na.rm=na.rm),
                se=sd(measurevar, na.rm=na.rm)/n(),
                ci=se*qt(conf.interval/2 + .5, n()-1))

    return(datac)
}
  
```

```{r, echo=FALSE} 
pd <- position_dodge(0.1) # move them .05 to the left and right

dat2%>%
  mutate(Age=ifelse(AgeGe40==1,"Age>=40","Age<40"))%>%
      select(pct_recruit, num_coupon, Age, contact_mode)%>%
      group_by(num_coupon, Age, contact_mode)%>%
      summarise(N=n(),
                pct_recruit=mean (pct_recruit, na.rm=T),
                sd=sd(pct_recruit, na.rm=T),
                se=sd(pct_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=pct_recruit, color=contact_mode))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
facet_grid(~Age)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.85, 0.2),
        legend.text=element_text(size=15),
        legend.title=element_blank())

```

```{r, echo=FALSE}
dat2%>%
  mutate(Age=ifelse(AgeGe40==1,"Age>=40","Age<40"))%>%
      select(any_recruit, num_coupon, Age, contact_mode)%>%
      group_by(num_coupon, Age, contact_mode)%>%
      summarise(N=n(),
                any_recruit=mean (any_recruit, na.rm=T),
                sd=sd(any_recruit, na.rm=T),
                se=sd(any_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=any_recruit, color=contact_mode))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
facet_grid(~Age)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.85, 0.2),
        legend.text=element_text(size=15),
        legend.title=element_blank())



#dat%>%
#  mutate(Age=ifelse(AgeGe40==1,"Age>=40","Age<40"))%>%
#ggplot(aes(x=num_coupon,y=num_recruit, color=Age))+
#facet_grid(~contact_mode)+
#  geom_smooth(method="lm", se=FALSE)

```

 
# 4. Coupon type
```{r, echo=FALSE}

dat2%>%
  group_by(coupon_type)%>%
  summarise(n=n(),
            m=mean(any_recruit),
            se = sd(any_recruit, na.rm = TRUE) / sqrt(sum(!is.na(any_recruit))))
``` 

```{r}
plot_coupon <- dat2 %>%
  group_by(coupon_type) %>%
  summarise(
    n = n(),
    m = mean(any_recruit, na.rm = TRUE),
    se = sd(any_recruit, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )
plot_coupon <- plot_coupon %>%
  mutate(
    coupon_label = paste0(coupon_type, "\n(n=", n, ")")
  )

```

```{r}
p_coupon<-ggplot(plot_coupon, aes(x = coupon_label, y = m)) +
  geom_col(fill = "#0072B2", color = "black", width = 0.5, linewidth = 0.4) +
  geom_errorbar(
    aes(ymin = m - 1.96*se, ymax = m + 1.96*se),
    width = 0.1, linewidth = 0.4, color = "black"
  ) +
  scale_y_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    x = "Coupon type",
    y = "Any recruits"
  ) +
  theme_minimal(base_size = 14)+
  theme(
     axis.title.x = element_text(size = 16, face = "bold"),
     axis.title.y = element_text(size = 16, face = "bold"),
     axis.text.x = element_text(size = 16),
     axis.text.y = element_text(size = 14),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
   )


ggsave(filename = "O:/Korean Health/Analysis/SL/NCSES/Figure_Coupon_NCSES.jpg",
       plot = p_coupon, width = 8, height = 5, dpi = 100, bg = "white")

```


- Wilcoxon Test 
```{r}
wilcox.test(num_recruit~coupon_type,dat2%>%filter(contact_mode=="Email"))
wilcox.test(num_recruit~coupon_type,dat2%>%filter(contact_mode=="SMS"))
wilcox.test(any_recruit~coupon_type,dat2%>%filter(contact_mode=="Email"))
wilcox.test(any_recruit~coupon_type,dat2%>%filter(contact_mode=="SMS"))
wilcox.test(pct_recruit~coupon_type,dat2%>%filter(contact_mode=="Email"))
wilcox.test(pct_recruit~coupon_type,dat2%>%filter(contact_mode=="SMS"))

table(dat2$contact_mode)
```

```{r, echo=FALSE}

dat2%>%
      select(any_recruit, num_coupon, coupon_type)%>%
      group_by(num_coupon, coupon_type)%>%
      summarise(N=n(),
                any_recruit=mean (any_recruit, na.rm=T),
                sd=sd(any_recruit, na.rm=T),
                se=sd(any_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=any_recruit, color=coupon_type))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.8, 0.23),
        legend.text=element_text(size=15),
        legend.title=element_blank())
```

```{r, echo=FALSE}
dat%>%
      select(pct_recruit, num_coupon, coupon_type)%>%
      group_by(num_coupon, coupon_type)%>%
      summarise(N=n(),
                percent_recruit=mean (pct_recruit, na.rm=T),
                sd=sd(pct_recruit, na.rm=T),
                se=sd(pct_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=percent_recruit, color=coupon_type))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.8, 0.23),
        legend.text=element_text(size=15),
        legend.title=element_blank())
```

```{r, echo=FALSE}
dat%>%
      select(num_recruit, num_coupon, coupon_type)%>%
      group_by(num_coupon, coupon_type)%>%
      summarise(N=n(),
                num_recruit=mean (num_recruit, na.rm=T),
                sd=sd(num_recruit, na.rm=T),
                se=sd(num_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=num_recruit, color=coupon_type))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.8, 0.23),
        legend.text=element_text(size=15),
        legend.title=element_blank())


```

## 4.1 Email


```{r, echo=FALSE}
dat%>%
  filter(contact_mode=="Email")%>%
  mutate(Age=ifelse(AgeGe40==1,"Age>=40","Age<40"))%>%
      select(any_recruit, num_coupon, Age, coupon_type)%>%
      group_by(num_coupon, Age, coupon_type)%>%
      summarise(N=n(),
                any_recruit=mean (any_recruit, na.rm=T),
                sd=sd(any_recruit, na.rm=T),
                se=sd(any_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=any_recruit, color=coupon_type))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
facet_grid(~Age)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.8, 0.2),
        legend.text=element_text(size=15),
        legend.title=element_blank())
```

```{r, echo=FALSE}
dat%>%
  filter(contact_mode=="Email")%>%
  mutate(Age=ifelse(AgeGe40==1,"Age>=40","Age<40"))%>%
      select(pct_recruit, num_coupon, Age, coupon_type)%>%
      group_by(num_coupon, Age, coupon_type)%>%
      summarise(N=n(),
                percent_recruit=mean (pct_recruit, na.rm=T),
                sd=sd(pct_recruit, na.rm=T),
                se=sd(pct_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=percent_recruit, color=coupon_type))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
facet_grid(~Age)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.8, 0.2),
        legend.text=element_text(size=15),
        legend.title=element_blank())
```

```{r, echo=FALSE}
dat%>%
  filter(contact_mode=="Email")%>%
  mutate(Age=ifelse(AgeGe40==1,"Age>=40","Age<40"))%>%
      select(num_recruit, num_coupon, Age, coupon_type)%>%
      group_by(num_coupon, Age, coupon_type)%>%
      summarise(N=n(),
                num_recruit=mean (num_recruit, na.rm=T),
                sd=sd(num_recruit, na.rm=T),
                se=sd(num_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=num_recruit, color=coupon_type))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
facet_grid(~Age)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.8, 0.2),
        legend.text=element_text(size=15),
        legend.title=element_blank())


datE<-dat%>%
  filter(contact_mode=="Email")

```

```{r}

wilcox.test(datE$any_recruit~datE$coupon_type, paired=FALSE)
wilcox.test(datE$pct_recruit~datE$coupon_type, paired=FALSE)
wilcox.test(datE$num_recruit~datE$coupon_type, paired=FALSE)

wilcox.test(any_recruit~coupon_type, paired=FALSE, data=subset(datE,AgeGe40==1))
wilcox.test(any_recruit~coupon_type, paired=FALSE, data=subset(datE,AgeGe40==0))
wilcox.test(pct_recruit~coupon_type, paired=FALSE, data=subset(datE,AgeGe40==1))
wilcox.test(pct_recruit~coupon_type, paired=FALSE, data=subset(datE,AgeGe40==0))
wilcox.test(num_recruit~coupon_type, paired=FALSE, data=subset(datE,AgeGe40==1))
wilcox.test(num_recruit~coupon_type, paired=FALSE, data=subset(datE,AgeGe40==0))

```

```{r, echo=FALSE,eval=FALSE}
 
hawk%>%
  group_by(coupon_type)%>%
  summarise(p=mean(pct_recruit,na.rm=T))

dat%>%
  group_by(coupon_type)%>%
  summarise(p=mean(pct_recruit))

ds<-summarySE(dat%>%
  mutate(Age=ifelse(AGEP>=30,"Age>=30","Age<30")), 
  measurevar="num_recruit", groupvars=c("Age","num_coupon","coupon_type"))

ds%>%
ggplot(aes(x=num_coupon,y=num_recruit, color=Age))+
  geom_line(position=pd)+
  geom_point(position=pd)+
facet_grid(~coupon_type)

dat2%>%filter(contact_mode=="Email")%>%
  tabyl(any_recruit,coupon_type)
 
dat2E<-dat2%>%filter(contact_mode=="Email")

summary(glm(any_recruit~coupon_type+num_coupon, 
            subset(dat2,contact_mode=="Email"),
            family=quasibinomial))


summary(glm(any_recruit~coupon_type+num_coupon, 
            subset(dat2,contact_mode=="SMS"),
            family=quasibinomial))


summary(glm(num_recruit~coupon_type+num_coupon, 
            subset(dat2,contact_mode=="Email"),
            family=quasipoisson))


summary(glm(num_recruit~coupon_type+num_coupon, 
            subset(dat2,contact_mode=="SMS"),
            family=quasipoisson))


summary(glm(num_recruit~coupon_type, dat2, family=quasipoisson()))


dat2E%>%
tabyl(num_recruit,coupon_type)%>%
adorn_percentages("col")%>%
  adorn_pct_formatting(digits=1)%>%
  adorn_ns()


dat2E%>%
tabyl(any_recruit,coupon_type)%>%
adorn_percentages("col")%>%
  adorn_pct_formatting(digits=1)%>%
  adorn_ns()



dat2E%>%
  group_by(coupon_type)%>%
  summarise(m_p=mean(pct_recruit),
            m_n=mean(num_recruit),
            m_a=mean(any_recruit))%>%
  summarise(p.value = t.test(pct_recruit ~ coupon_type, paired = TRUE)$p.value)

dat2E %>%
  group_by(coupon_type) %>%
  mutate(pval = wilcox.test(pct_recruit~coupon_type, paired=FALSE)$p.value)


wilcox.test(dat2E$pct_recruit~dat2E$coupon_type, paired=FALSE)
wilcox.test(dat2E$num_recruit~dat2E$coupon_type, paired=FALSE)
wilcox.test(dat2E$any_recruit~dat2E$coupon_type, paired=FALSE)

dat%>%
  filter(contact_mode=="Email")%>%
  group_by(coupon_type)%>%
  summarise(p=mean(pct_recruit))

datE<-dat%>%
  filter(contact_mode=="Email")
wilcox.test(datE$pct_recruit~datE$coupon_type, paired=FALSE)
wilcox.test(datE$num_recruit~datE$coupon_type, paired=FALSE)
wilcox.test(datE$any_recruit~datE$coupon_type, paired=FALSE)

dat2E%>%
  group_by(coupon_type)%>%
  summarise(m=mean(pct_recruit))

dat2E%>%
  group_by(coupon_type)%>%
  summarise(m=mean(pct_recruit))

 

dat2E%>%
  tabyl(coupon_type) 

hawk%>%
  filter(contact_mode=="Email")%>%
  mutate(seed=ifelse(wave==0,"1.Seed","2.NonSeed"))%>%
  tabyl(coupon_type,seed)%>%
  adorn_percentages("col")%>%
  adorn_pct_formatting(digits=1)%>%
  adorn_ns()

View(seed)

table(hawk$coupon_type)

table(is.na(hawk$seed_id),hawk$wave,exclude=NULL)

View(hawk%>%
group_by(seed_id)%>%
  select(seed_id,coupon_type))


sum(hawk$coupon_type=="E-All")

CouponType<-hawk%>%
group_by(seed_id)%>%
  summarise(p_EA=sum(coupon_type=="E-All")/n(),
            p_ES=sum(coupon_type=="E-Sep")/n(),
            p_SI=sum(coupon_type=="S-Img")/n(),
            p_ST=sum(coupon_type=="S-Text")/n(),
            ChainN=n())%>%
  left_join(ChainLength,by="seed_id")
#View(CouponType)

CouponType%>%
  select(p_EA,p_ES,p_ST,p_SI,ChainLength,ChainN)%>%
  cor()
install.packages("psych")
library(psych)
paired.r(cor(CouponType$p_EA,CouponType$ChainN), 
         cor(CouponType$p_ES,CouponType$ChainN), 
         n=56)

paired.r(cor(CouponType$p_SI,CouponType$ChainN), 
         cor(CouponType$p_ST,CouponType$ChainN), 
         n=56)


table(hawk$coupon_type=="E-All")[2]/table(!is.na(hawk$coupon_type))[1]

hawk%>%
  filter(contact_mode=="SMS")%>%
  mutate(seed=ifelse(wave==0,"1.Seed","2.NonSeed"))%>%
  tabyl(coupon_type,seed)%>%
  adorn_percentages("col")%>%
  adorn_pct_formatting(digits=1)%>%
  adorn_ns()

```



## 5.2 Text 

```{r, echo=FALSE}
dat%>%
  filter(contact_mode=="SMS")%>%
  mutate(Age=ifelse(AgeGe40==1,"Age>=40","Age<40"))%>%
      select(any_recruit, num_coupon, Age, coupon_type)%>%
      group_by(num_coupon, Age, coupon_type)%>%
      summarise(N=n(),
                any_recruit=mean (any_recruit, na.rm=T),
                sd=sd(any_recruit, na.rm=T),
                se=sd(any_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=any_recruit, color=coupon_type))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
facet_grid(~Age)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.15, 0.8),
        legend.text=element_text(size=15),
        legend.title=element_blank())
```

```{r, echo=FALSE}
dat%>%
  filter(contact_mode=="SMS")%>%
  mutate(Age=ifelse(AgeGe40==1,"Age>=40","Age<40"))%>%
      select(pct_recruit, num_coupon, Age, coupon_type)%>%
      group_by(num_coupon, Age, coupon_type)%>%
      summarise(N=n(),
                percent_recruit=mean (pct_recruit, na.rm=T),
                sd=sd(pct_recruit, na.rm=T),
                se=sd(pct_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=percent_recruit, color=coupon_type))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
facet_grid(~Age)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.15, 0.8),
        legend.text=element_text(size=15),
        legend.title=element_blank())

```

```{r, echo=FALSE}
dat%>%
  filter(contact_mode=="SMS")%>%
  mutate(Age=ifelse(AgeGe40==1,"Age>=40","Age<40"))%>%
      select(num_recruit, num_coupon, Age, coupon_type)%>%
      group_by(num_coupon, Age, coupon_type)%>%
      summarise(N=n(),
                num_recruit=mean (num_recruit, na.rm=T),
                sd=sd(num_recruit, na.rm=T),
                se=sd(num_recruit, na.rm=T)/n(),
                ci=se*qt(.95/2 + .5, n()-1))%>%
ggplot(aes(x=num_coupon,y=num_recruit, color=coupon_type))+
  geom_line(position=pd, size=2)+
  geom_point(position=pd)+
facet_grid(~Age)+
  theme(strip.text.x = element_text(size = 18),
        axis.title = element_text(size = 15),
        legend.position = c(0.15, 0.8),
        legend.text=element_text(size=15),
        legend.title=element_blank())

datS<-dat%>%
  filter(contact_mode=="SMS")
```

```{r}
wilcox.test(datS$pct_recruit~datS$coupon_type, paired=FALSE)
wilcox.test(datS$num_recruit~datS$coupon_type, paired=FALSE)
wilcox.test(datS$any_recruit~datS$coupon_type, paired=FALSE)

wilcox.test(any_recruit~coupon_type, paired=FALSE, data=subset(datS,AgeGe40==1))
wilcox.test(any_recruit~coupon_type, paired=FALSE, data=subset(datS,AgeGe40==0))
wilcox.test(pct_recruit~coupon_type, paired=FALSE, data=subset(datS,AgeGe40==1))
wilcox.test(pct_recruit~coupon_type, paired=FALSE, data=subset(datS,AgeGe40==0))
wilcox.test(num_recruit~coupon_type, paired=FALSE, data=subset(datS,AgeGe40==1))
wilcox.test(num_recruit~coupon_type, paired=FALSE, data=subset(datS,AgeGe40==0))

```