---
title: "PCA (PATH 2017)-10.05.2026"
output: html_document
date: "2026-05-10"
---

# Set up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, include=FALSE}
if (!require("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(
  dplyr, tidyr, ggplot2, ggforce, stringr, forcats, purrr,
  janitor, skimr, gtsummary, writexl, vcd, reshape2, haven
)
```

```{r}
PATH2017 <- read_sas("O:\\slteam\\RDS\\Study 2_IDU\\Data\\AO\\merged_data.sas7bdat")
dim(PATH2017)
df <- PATH2017

df$SNW

df %>% 
  select(M_SNQ2, M_SNQ1_1, M_SNQ1_2, M_SNQ2, M_NETSIZE)
```

```{r}
# summarise_col <- function(df, col) {
#   x <- df[[col]]
#   n_total <- length(x)
#   n_na    <- sum(is.na(x))
#   n_valid <- n_total - n_na
# 
#   cat("Column:", col, "\n")
#   cat("Total:", n_total, "| Valid:", n_valid, "| NAs:", n_na, "\n\n")
# 
#   if (is.numeric(x)) {
#     # ── Numeric ──────────────────────────────────────────
#     cat("Type: numeric\n")
#     cat(sprintf(
#       "  Min    : %g\n  Q1     : %g\n  Median : %g\n  Mean   : %g\n  Q3     : %g\n  Max    : %g\n  SD     : %g\n",
#       min(x, na.rm = TRUE),
#       quantile(x, 0.25, na.rm = TRUE),
#       median(x, na.rm = TRUE),
#       mean(x, na.rm = TRUE),
#       quantile(x, 0.75, na.rm = TRUE),
#       max(x, na.rm = TRUE),
#       sd(x, na.rm = TRUE)
#     ))
# 
#   } else {
#     # ── Categorical (factor, character, logical) ──────────
#     cat("Type: categorical\n\n")
#     tbl  <- sort(table(x, useNA = "no"), decreasing = TRUE)
#     pcts <- round(100 * tbl / n_valid, 1)
# 
#     df_out <- data.frame(
#       label   = names(tbl),
#       n       = as.integer(tbl),
#       percent = paste0(pcts, "%"),
#       stringsAsFactors = FALSE
#     )
#     print(df_out, row.names = FALSE)
#     cat("\nLevels:", length(tbl), "\n")
#   }
# }
# summary_stats <- function(df, col) {
#   x <- df[[col]]
#   
#   list(
#     n       = length(x),
#     n_na    = sum(is.na(x)),
#     mean    = mean(x, na.rm = TRUE),
#     sd      = sd(x, na.rm = TRUE),
#     min     = min(x, na.rm = TRUE),
#     q1      = quantile(x, 0.25, na.rm = TRUE),
#     median  = median(x, na.rm = TRUE),
#     q3      = quantile(x, 0.75, na.rm = TRUE),
#     max     = max(x, na.rm = TRUE)
#   )
# }
```

```{r}
# col <- "M_HQ3"
# 
# summarise_col(df, col)
# unique(df[[col]])
# df$IS_SC5_OTHER
```

# Preprocessing
## Get required variables
```{r}
var_list <- c(
  # grouping
  "WAVE_NUM", # wave number
  "SEED", # is seed or not
  
  # sociodemographic
  "PS_Q5", # age
  "M_AQ1", # gender
  grep("^M_AQ3_", names(df), value = TRUE), # race
  "M_AQ6", # education
  "M_HQ1", # employment stats
  "M_AQ4", # marital status
  "M_HQ3", # income
  "PS_Q3", # lives in greater detroit
  "M_HQ9", # homeless in the past year
  "M_HQ9_1", # homeless currently
  
  # injection behaviour
  grep("^IS_SC5_", names(df), value = TRUE), # drugs used
  "IS_SC6", # drug injected most often
  grep("^M_DQ1_", names(df), value = TRUE), # age when first using IS_SC6
  "M_DQ3", # Frequency of use for drug injected most often in the past 12 months
  "M_DQ7", # frequency of use of sterile needle in past 12 months
  "M_DQ28", # ever overdosed on a drug
  "M_DQ11", # need to use more [IS_SC6] than in the past in past 12 months
  "M_DQ12", # past 12 Months: Want to cut down or stop [IS_SC6]
  "M_DQ13", # has problems with emotions, nerves, mental health as result of use of [IS_SC6] in past 12 months
  "M_DQ14", # has physical health problems as result of use of [IS_SC6] in past 12 months
  "M_DQ15", # Problems with family as result of use of [SC6] in past 12 months
  "M_DQ16", # Problems at work, home, school as result of use of [SC6] in past 12 months
  "M_DQ17", # Has trouble with the law due to injecting drugs in past 12 months
  "M_DQ28_3", # Ever received Narcan
  "M_DQ31", # Respondent, respondent's family/friends have access to Narcan
  "M_NUMBER_INT", #	Network size: Known injection drug users interact with more than once a week
  
  # mental health
  "M_EQ1", # global life satisfaction
  
  # physical health
  "M_BQ1", # self-rate health
  "M_GQ5", # ever had HIV test
  "M_GQ8", # ever HIV positive
  "M_BQ2", # Ever tested for hepatitis C
  "M_BQ3", # Respondent ever has hepatitis C
  "M_BQ10", # Feeling age (years)
  
  # social network
  "M_NUMBER_CLS", # Known injection drug users respondent feel close to
  "M_NUMBER_CLS", # Known injection drug users respondent feel close to
  "M_NETSIZE", # Computed, people known who inject
  "M_NUMBER_INT" # 	Network size: Known injection drug users interact with more than once a week
)

var_list_exist <- var_list[var_list %in% names(df)]
missing_cols <- setdiff(var_list, var_list_exist)

if(length(missing_cols) > 0){
  message("These variables are not found in df: ", paste(missing_cols, collapse = ", "))
}

dat_raw <- df %>%
  dplyr::select(all_of(var_list_exist))

# Drop instances where SEED is empty
dat_raw <- dat_raw %>% 
  filter(!is.na(SEED))
```

## Create binary variables

```{r}
dat_raw %>%
  group_by(IS_SC6) %>%
  tally()
```

```{r}
dat_raw %>%
  filter(IS_SC5_OS!="") %>%
  select(IS_SC5_OS)
# IS_SC5 allows for multiselect
# IS_SC6 is only asked if respondents choose more than one drug in IS_SC5

test <- dat_raw %>%
  dplyr::mutate(
    multi.drug = case_when(IS_SC5_SPDBALL + IS_SC5_HEROIN + IS_SC5_COCAINE + IS_SC5_CRACK + IS_SC5_CRYSTAL + IS_SC5_OPIATE + IS_SC5_OTHER > 1 ~ 1,
                           IS_SC5_SPDBALL + IS_SC5_HEROIN + IS_SC5_COCAINE + IS_SC5_CRACK + IS_SC5_CRYSTAL + IS_SC5_OPIATE + IS_SC5_OTHER <= 1 ~ 0,
                           TRUE ~ NA_real_),
    heroin_most_freq = case_when(IS_SC5_HEROIN == 1 & multi.drug == 0 ~ 1,               # Respondent only uses heroin
                                 multi.drug == 1 & IS_SC6 == 2 ~ 1,                      # Respondent uses multiple drugs and heroin IS the most frequently used
                                 IS_SC5_HEROIN == 0 ~ 0,                                 # Respondent doesn't report using heroin
                                 multi.drug == 1 & IS_SC6 != 2 ~ 0,                      # Respondent uses multiple drugs and heroin is not the most frequently used
                                 TRUE ~ NA_real_),
    # Get the age the respondent was when they first started injecting their most frequently used drug
    DQ3 = case_when((IS_SC5_SPDBALL == 1 & multi.drug == 0) |  (multi.drug == 1 & IS_SC6 == 1) ~ M_DQ1_1,
                    (IS_SC5_HEROIN == 1 & multi.drug == 0) | (multi.drug == 1 & IS_SC6 == 2) ~ M_DQ1_2,
                    (IS_SC5_COCAINE == 1 & multi.drug == 0) | (multi.drug == 1 & IS_SC6 == 3) ~ M_DQ1_3,
                    (IS_SC5_CRACK == 1 & multi.drug == 0) | (multi.drug == 1 & IS_SC6 == 4) ~ M_DQ1_4,
                    (IS_SC5_CRYSTAL == 1 & multi.drug == 0) | (multi.drug == 1 & IS_SC6 == 5) ~ M_DQ1_5,
                    (IS_SC5_OPIATE == 1 & multi.drug == 0) ~ M_DQ1_6, # IC_SC6 doesn't have a level corresponding to 'opiate'
                    (IS_SC5_OTHER == 1 & multi.drug == 0) ~ M_DQ1_7, # IC_SC6 doesn't have a level corresponding to 'other'
                    TRUE ~ NA_real_ ),
    INJECT_NUM = IS_SC5_SPDBALL + IS_SC5_HEROIN + IS_SC5_COCAINE + IS_SC5_CRACK + IS_SC5_CRYSTAL + IS_SC5_OPIATE + IS_SC5_OTHER,
  )
test %>%
  group_by(multi.drug) %>%
  tally()
test %>%
  group_by(heroin_most_freq) %>%
  tally()
test %>%
  select(IS_SC5_SPDBALL , IS_SC5_HEROIN , IS_SC5_COCAINE , IS_SC5_CRACK , IS_SC5_CRYSTAL , IS_SC5_OPIATE , IS_SC5_OTHER, IS_SC6, INJECT_NUM, multi.drug, heroin_most_freq,
         M_DQ1_1, M_DQ1_2,  M_DQ1_3,  M_DQ1_4,  M_DQ1_5,  M_DQ1_6,  M_DQ1_7, DQ3)
test %>%
  select(M_DQ1_1, M_DQ1_2,  M_DQ1_3,  M_DQ1_4,  M_DQ1_5,  M_DQ1_6,  M_DQ1_7)
```

```{r}
income_cut <- 10000
inject_years_cut <- 15

dat_bin <- dat_raw %>%
  dplyr::mutate(
    seed01 = dplyr::case_when(
      SEED == 1 ~ 1,
      SEED == 0 ~ 0,
      TRUE ~ NA_real_
    ),
    seed_group = factor(seed01, levels = c(0, 1), labels = c("Non-seed", "Seed")),
    
    ## helper continuous variables
    # Calculate the difference between how someone feels and how hold they actually are
    # Introduce difference upper limit of 100 -> in these cases set age_feel_diff to 0
    age_feel_diff = dplyr::case_when(
      !is.na(M_BQ10) & !is.na(PS_Q5) & abs(M_BQ10 - PS_Q5) <= 100 ~ M_BQ10 - PS_Q5,
      TRUE ~ NA_real_
    ),
    
    multi.drug = case_when(IS_SC5_SPDBALL + IS_SC5_HEROIN + IS_SC5_COCAINE + IS_SC5_CRACK + IS_SC5_CRYSTAL + IS_SC5_OPIATE + IS_SC5_OTHER > 1 ~ 1,
                           IS_SC5_SPDBALL + IS_SC5_HEROIN + IS_SC5_COCAINE + IS_SC5_CRACK + IS_SC5_CRYSTAL + IS_SC5_OPIATE + IS_SC5_OTHER <= 1 ~ 0,
                           TRUE ~ NA_real_),
    
    # years of injecting (current age less age when first injected)
    DQ3 = case_when((IS_SC5_SPDBALL == 1 & multi.drug == 0) |  (multi.drug == 1 & IS_SC6 == 1) ~ M_DQ1_1,
                    (IS_SC5_HEROIN == 1 & multi.drug == 0) | (multi.drug == 1 & IS_SC6 == 2) ~ M_DQ1_2,
                    (IS_SC5_COCAINE == 1 & multi.drug == 0) | (multi.drug == 1 & IS_SC6 == 3) ~ M_DQ1_3,
                    (IS_SC5_CRACK == 1 & multi.drug == 0) | (multi.drug == 1 & IS_SC6 == 4) ~ M_DQ1_4,
                    (IS_SC5_CRYSTAL == 1 & multi.drug == 0) | (multi.drug == 1 & IS_SC6 == 5) ~ M_DQ1_5,
                    (IS_SC5_OPIATE == 1 & multi.drug == 0) ~ M_DQ1_6, # IC_SC6 doesn't have a level corresponding to 'opiate'
                    (IS_SC5_OTHER == 1 & multi.drug == 0) ~ M_DQ1_7, # IC_SC6 doesn't have a level corresponding to 'other'
                    TRUE ~ NA_real_ ),
    inject_years = dplyr::case_when(
      !is.na(DQ3) & !is.na(PS_Q5) & DQ3 < PS_Q5 ~ PS_Q5 - DQ3,
      TRUE ~ NA_real_
    ),
    
    # Sociodemographic
    age_lt40 = dplyr::case_when(
      PS_Q5 < 40 ~ 1,
      !is.na(PS_Q5) ~ 0,
      TRUE ~ NA_real_
    ),
    
    male = dplyr::case_when(
      M_AQ1 == 1 ~ 1, # males == 1
      M_AQ1 %in% c(0, 2) ~ 0, # females == 0
      TRUE ~ NA_real_
    ),
    
    race_nh_black = dplyr::case_when(
      M_AQ3_BLACK == 1 ~ 1,
      M_AQ3_AIAN == 1 & M_AQ3_BLACK != 1 ~ 0,
      M_AQ3_ASIAN == 1 & M_AQ3_BLACK != 1  ~ 0,
      M_AQ3_MENA == 1 & M_AQ3_BLACK != 1  ~ 0,
      M_AQ3_NHPI == 1 & M_AQ3_BLACK != 1  ~ 0,
      M_AQ3_WHITE == 1 & M_AQ3_BLACK != 1  ~ 0,
      M_AQ3_WHITE == 1 & M_AQ3_BLACK != 1  ~ 0,  
      TRUE ~ NA_real_
    ),
    
    somecollege_plus = dplyr::case_when(
      M_AQ6 >= 6 ~ 1,
      !is.na(M_AQ6) ~ 0,
      TRUE ~ NA_real_
    ),
    
    employed = dplyr::case_when(
      M_HQ1 %in% c(1, 2) ~ 1, 
      M_HQ1 %in% c(4, 5, 6, 7) ~ 0,
      TRUE ~ NA_real_
    ),
    
    married_partnered = dplyr::case_when(
      M_AQ4 %in% c(1, 2) ~ 1,
      M_AQ4 %in% c(3, 4, 5) ~ 0,
      TRUE ~ NA_real_
    ),

    income_le10000 = dplyr::case_when(
      M_HQ3 >= income_cut ~ 1,
      !is.na(M_HQ3) ~ 0,
      TRUE ~ NA_real_
    ),
    
    homeless_any = dplyr::case_when(
      M_HQ9 == 1 | M_HQ9_1 == 1 ~ 1,
      M_HQ9 %in% c(0, 2) | M_HQ9_1 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # Injection Behavior
    multi.drug = case_when(IS_SC5_SPDBALL + IS_SC5_HEROIN + IS_SC5_COCAINE + IS_SC5_CRACK + IS_SC5_CRYSTAL + IS_SC5_OPIATE + IS_SC5_OTHER > 1 ~ 1,
                           IS_SC5_SPDBALL + IS_SC5_HEROIN + IS_SC5_COCAINE + IS_SC5_CRACK + IS_SC5_CRYSTAL + IS_SC5_OPIATE + IS_SC5_OTHER <= 1 ~ 0,
                           TRUE ~ NA_real_),
    heroin_most_freq = case_when(IS_SC5_HEROIN == 1 & multi.drug == 0 ~ 1,               # Respondent only uses heroin
                                 multi.drug == 1 & IS_SC6 == 2 ~ 1,                      # Respondent uses multiple drugs and heroin is the most frequently used
                                 IS_SC5_HEROIN == 0 ~ 0,                                 # Respondent doesn't report using heroin
                                 multi.drug == 1 & IS_SC6 != 2 ~ 0,                      # Respondent uses multiple drugs and heroin is not the most frequently used
                                 TRUE ~ NA_real_),
    
    INJECT_NUM = IS_SC5_SPDBALL + IS_SC5_HEROIN + IS_SC5_COCAINE + IS_SC5_CRACK + IS_SC5_CRYSTAL + IS_SC5_OPIATE + IS_SC5_OTHER,
    
    drugs_injected_ge2 = dplyr::case_when(
      INJECT_NUM >= 2 ~ 1,
      !is.na(INJECT_NUM) ~ 0,
      TRUE ~ NA_real_
    ),
    
    inject_timelength = dplyr::case_when(
      !is.na(inject_years) & inject_years >= inject_years_cut ~ 1,
      !is.na(inject_years) ~ 0,
      TRUE ~ NA_real_
    ),
    
    inject_more_than_once_day = dplyr::case_when(
      M_DQ3 == 1 ~ 1,
      !is.na(M_DQ3) ~ 0,
      TRUE ~ NA_real_
    ),
    
    sterile_needle_always = dplyr::case_when(
      M_DQ7 == 5 ~ 1,
      !is.na(M_DQ7) ~ 0,
      TRUE ~ NA_real_
    ),
    
    need_more_drug_same_effect = dplyr::case_when(
      M_DQ11 == 1 ~ 1,
      M_DQ11 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    tried_reduce_stop_drug = dplyr::case_when(
      M_DQ12 == 1 ~ 1,
      M_DQ12 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    drug_emotional_physical_issue = dplyr::case_when(
      M_DQ13 == 1  ~ 1,
      M_DQ13 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),

    drug_family_issue = dplyr::case_when(
      M_DQ15 == 1 ~ 1,
      M_DQ15 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    drug_homeworkschool_issue = dplyr::case_when(
      M_DQ16 == 1 ~ 1,
      M_DQ16 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    drug_legal_issue = dplyr::case_when(
      M_DQ17 == 1 ~ 1,
      M_DQ17 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    ever_overdose = dplyr::case_when(
      M_DQ28 == 1 ~ 1,
      M_DQ28 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),

    have_narcan = dplyr::case_when(
      M_DQ28_3 == 1 ~ 1,
      M_DQ28_3 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),

    narcan_access = dplyr::case_when(
      M_DQ31 == 1 ~ 1,
      M_DQ31 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # Mental health
    life_satisfaction_low = dplyr::case_when(
      M_EQ1 %in% c(4, 5) ~ 1,
      M_EQ1 %in% c(1, 2, 3) ~ 0,
      TRUE ~ NA_real_
    ),
    
    age_feel_older = dplyr::case_when(
      !is.na(age_feel_diff) & age_feel_diff > 0 ~ 1,
      !is.na(age_feel_diff) & age_feel_diff <= 0 ~ 0,
      TRUE ~ NA_real_
    ), 
    
    # Physical health
    srh_fairpoor = dplyr::case_when(
      M_BQ1 %in% c(4, 5) ~ 1,
      M_BQ1 %in% c(1, 2, 3) ~ 0,
      TRUE ~ NA_real_
    ),
    
    hiv_tested = dplyr::case_when(
      M_GQ5 == 1 ~ 1,
      M_GQ5 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    hiv_positive = dplyr::case_when(
      M_GQ5 == 1 & M_GQ8 == 1 ~ 1,
      M_GQ5 == 1 & M_GQ8 %in% c(0, 2) ~ 0,
      M_GQ5 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    hepc_tested = dplyr::case_when(
      M_BQ2 == 1 ~ 1,
      M_BQ2 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),

    hepc_diagnosed = dplyr::case_when(
      M_BQ2 == 1 & M_BQ3 == 1 ~ 1,
      M_BQ2 == 1 & M_BQ3 %in% c(0, 2) ~ 0,
      M_BQ2 %in% c(0, 2) ~ 0,
      TRUE ~ NA_real_
    ),
    
    # Social Network
    pwid_network_gt10 = dplyr::case_when(
      M_NETSIZE > 10 ~ 1,
      !is.na(M_NETSIZE) ~ 0,
      TRUE ~ NA_real_
    ),
    
    interact_pwid_weekly_any = dplyr::case_when(
      M_NUMBER_INT > 0 ~ 1,
      !is.na(M_NUMBER_INT) ~ 0,
      TRUE ~ NA_real_
    )
  ) 

dat_raw %>% # oh wow really no missing
  filter(is.na(M_GQ5) | M_GQ5=="") %>% 
  select(M_GQ5)
dat_raw %>% # oh wow there are a lot of missing
  filter(is.na(IS_SC6) | IS_SC6=="") %>% 
  select(IS_SC6)

dat_bin %>%
  group_by(heroin_most_freq) %>%
  tally()
```

## Binary data checks
```{r}
dat_bin %>%
  group_by(heroin_most_freq) %>%
  tally() %>%
  mutate(perc = round(n / sum(n) * 100, 1))

vars <- c("DQ3")

summary_table <- purrr::map_dfr(vars, function(var) {
  col <- dat_bin[[var]]
  unique_vals <- sort(unique(col[!is.na(col)]))
  
  tibble(
    variable      = var,
    unique_values = paste(unique_vals, collapse = ", "),
    n_unique      = length(unique_vals),
    n_missing     = sum(is.na(col)),
    pct_missing   = round(mean(is.na(col)) * 100, 1)
  )
})

print(summary_table)

dat_bin %>%
  group_by(INJECT_NUM) %>%
  tally() %>%
  mutate(perc = round(n / sum(n) * 100, 1))
```

## Correlation test

```{r}
# Sociodemographic
cor(dat_bin[, c("homeless_any", "somecollege_plus", "income_le10000", "employed")], use = "pairwise.complete.obs")
```

```{r}
# Injection Behavior
cor(dat_bin[, c("drugs_injected_ge2", "heroin_most_freq", "inject_more_than_once_day", "inject_years", "sterile_needle_always", "need_more_drug_same_effect", "tried_reduce_stop_drug", "drug_emotional_physical_issue", "drug_family_issue", "drug_homeworkschool_issue", "drug_legal_issue", "ever_overdose", "have_narcan", "narcan_access")], use = "pairwise.complete.obs")
```

```{r}
# Mental Health - only one var, no need to run correlation test
# cor(dat_bin[, c("life_satisfaction_low")], use = "pairwise.complete.obs")
```

```{r}
# Physical Health
cor(dat_bin[, c("srh_fairpoor", "hiv_tested", "hiv_positive", "hepc_tested", "hepc_diagnosed", "age_feel_older")], use = "pairwise.complete.obs")
```

```{r}
# Social Network
cor(dat_bin[, c("interact_pwid_weekly_any", "pwid_network_gt10")], use = "pairwise.complete.obs")
```


# Missingness

```{r}
dat_miss_bin <- dat_bin %>%
  dplyr::select(
    seed01, seed_group,
    # Sociodemographic
    age_lt40,
    male,
    race_nh_black,
    married_partnered,
    income_le10000,
    
    # Mental Health
    life_satisfaction_low,
    
    # Physical Health
    srh_fairpoor,
    hiv_tested,
    
    # Injection Behavior
    drugs_injected_ge2,
    heroin_most_freq,
    inject_more_than_once_day,
    need_more_drug_same_effect,
    ever_overdose,
    narcan_access
  )

var_labels_bin <- c(
  # Sociodemographic
  age_lt40 = "age < 40",
  male = "male",
  race_nh_black = "non-Hispanic Black",
  married_partnered = "married or living with partner",
  income_le10000 = "income >= 10000",
  
  # Mental Health
  life_satisfaction_low = "life satisfaction low",
  
  # Physical Health
  srh_fairpoor = "self-rated health fair/poor",
  hiv_tested = "ever tested for HIV",
  
  # Injection Behavior
  drugs_injected_ge2 = "number of drugs injected >= 2",
  heroin_most_freq = "heroin most frequently used drug",
  inject_more_than_once_day = "inject > once/day",
  need_more_drug_same_effect = "need more drug for wanted effect",
  ever_overdose = "ever overdosed",
  narcan_access = "access to Narcan/Naloxone"
)

```

## Plot
```{r}
analysis_vars_bin <- setdiff(names(dat_miss_bin), c("seed01","seed_group"))

missing_overall_bin <- dat_miss_bin %>%
  dplyr::summarise(across(all_of(analysis_vars_bin), ~mean(is.na(.))*100)) %>%
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  dplyr::mutate(variable_label = dplyr::recode(variable, !!!var_labels_bin)) %>%
  dplyr::arrange(desc(pct_missing))

ggplot(missing_overall_bin,
       aes(x = reorder(variable_label, pct_missing), y = pct_missing)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "% Missing",
    title = "Missingness by binary variables (overall)"
  ) +
  theme_minimal()
```
## Table (for imputation understanding)
```{r}
missing_table_bin <- missing_overall_bin %>%
  dplyr::select(variable, variable_label, pct_missing) %>%
  dplyr::mutate(
    pct_missing = round(pct_missing, 2)
  ) %>%
  dplyr::arrange(desc(pct_missing))

print(missing_table_bin, n=11)

```

# PCA with imputation
## PCA variable

```{r}
## seed01 could be predictor due to it don't have missing value, but not putting in PCA main analysis
pca_vars <- setdiff(names(dat_miss_bin), c("CHAIN_WAVE", "seed01", "seed_group"))
pca_vars
imp_vars <- setdiff(names(dat_miss_bin), c("CHAIN_WAVE", "seed_group"))

## missing rate
miss_rate <- sapply(dat_miss_bin[imp_vars], function(x) mean(is.na(x)) * 100)

miss_df <- data.frame(
  variable = names(miss_rate),
  pct_missing = as.numeric(miss_rate)
) %>%
  arrange(pct_missing) %>% 
  filter(variable!="heroin_most_freq") 

miss_df
```

## Hot deck
```{r}
## one-variable hot deck imputation
# =============================================================================
# HOT DECK IMPUTATION (single variable)
#
# Hot deck imputation fills in missing values by "borrowing" observed values
# from similar records (donors). For each record with a missing target value
# (recipient), we find the k most similar complete records using Euclidean
# distance across the predictor columns, then randomly pick one of those k
# donors to supply the missing value.
#
# Arguments:
#   data       - a data frame that may contain NAs in the target column
#   target     - string name of the column to impute
#   predictors - character vector of column names used to measure similarity
#   k          - how many nearest neighbours to draw from (default 5)
#   seed       - random seed for reproducibility (default 123)
#
# Returns: a copy of `data` with NAs in `target` filled in.
# =============================================================================
hotdeck_one_var <- function(data, target, predictors, k = 5, seed = 123) {
  set.seed(seed)

  x <- data # Create copy

  miss_idx <- which(is.na(x[[target]])) # row positions missing target
  obs_idx  <- which(!is.na(x[[target]])) # row positions not missing target

  if (length(miss_idx) == 0) return(x) # if nothing is missing, do nothing
  
  # edge case: no predictors -> impute by random sampling from observed values (no distance matching)
  if (length(predictors) == 0) {
    x[[target]][miss_idx] <- sample(x[[target]][obs_idx], # draw from observed target values
                                    size = length(miss_idx), # one draw per missing 
                                    replace = TRUE) # allow the same donor more than once
    return(x)
  }
  # build predictor matrices for donors and recipients
  donor_mat <- as.matrix(x[obs_idx, predictors, drop = FALSE]) # predictors for complete rows
  recip_mat <- as.matrix(x[miss_idx, predictors, drop = FALSE]) # predictors for missing rows

  ## coerce all values to be type numeric
  donor_mat <- apply(donor_mat, 2, as.numeric)
  recip_mat <- apply(recip_mat, 2, as.numeric)

  ## If there is only one predictor, put in matrix
  if (is.null(dim(donor_mat))) donor_mat <- matrix(donor_mat, ncol = 1)
  if (is.null(dim(recip_mat))) recip_mat <- matrix(recip_mat, ncol = 1)

  donor_y <- x[[target]][obs_idx] # observed target values that can be donated
  
  # For every recipient row, find its k nearest donors and pick one
  imputed_values <- sapply(seq_len(nrow(recip_mat)), function(i) {
    # Get the squared euclidean distance from recipient i to every donor row
    d <- rowSums((t(t(donor_mat) - recip_mat[i, ]))^2)

    ## closest donor k
    k_use <- min(k, length(d)) # note; can't use more neighbours than donors available
    nearest_idx <- order(d)[1:k_use] # indices of the k donors with the smallest distance

    sample(donor_y[nearest_idx], size = 1) # randomly pick one of those k donors target values
  })

  x[[target]][miss_idx] <- imputed_values # add imputed values to dataframe
  x
}
```

## sequential hot deck from lowest missing rate
```{r}
dat_imp_seq <- dat_miss_bin

## let 0% missing be in the predictor pool
predictor_pool <- miss_df %>%
  filter(pct_missing == 0) %>%
  pull(variable)

## variables for sequentially impute
target_order <- miss_df %>%
  filter(pct_missing > 0) %>%
  arrange(pct_missing) %>%
  pull(variable)

predictor_pool
target_order
```

```{r}
## sequential hot deck
imputation_log <- data.frame(
  step = integer(),
  target = character(),
  pct_missing = numeric(),
  n_predictors_used = integer(),
  predictors_used = character()
)

for (j in seq_along(target_order)) {
  target_j <- target_order[j]

  ## predictors couldn't include themselves
  current_predictors <- setdiff(predictor_pool, target_j)

  dat_imp_seq <- hotdeck_one_var(
    data = dat_imp_seq,
    target = target_j,
    predictors = current_predictors,
    k = 5,
    seed = 100 + j
  )

  ## adding target_j in the predictor pool after imputing
  predictor_pool <- c(predictor_pool, target_j)

  imputation_log <- bind_rows(
    imputation_log,
    data.frame(
      step = j,
      target = target_j,
      pct_missing = miss_df$pct_missing[match(target_j, miss_df$variable)],
      n_predictors_used = length(current_predictors),
      predictors_used = paste(current_predictors, collapse = ", ")
    )
  )
}

imputation_log
```

## Check
```{r}
colSums(is.na(dat_imp_seq))
sum(is.na(dat_imp_seq[pca_vars]))
```

## PCA
```{r}
## imputed data for PCA
dat_pca_imp <- dat_imp_seq %>%
  dplyr::select(all_of(pca_vars))

dim(dat_pca_imp) # 905 x 17 (905 respondents x 17 attributes)

# subtracts the mean from each variable so everything is centered at 0
# divides by the standard deviation so all 17 variables are on the same scale
pca_imp <- prcomp(dat_pca_imp, center = TRUE, scale. = TRUE)

# Recall:
# - Standard deviation = sqrt(eigenval_i)
# - Proportion of Variance = eigenval_i / sum(all eigenval)
summary(pca_imp)
```
# PCA with imputation

```{r}
round(pca_imp$rotation,2)
```

```{r}
rot <- round(pca_imp$rotation, 2)[, 1:7]

lapply(colnames(rot), function(pc) {
  rot[abs(rot[, pc]) > 0.3, pc, drop = FALSE]
})
```

```{r}
# PC scores (each row = individual, each column = original data matrix dot.product eigenvector)
# pca_imp$x is the 905 x 17 matrix of PC scores from the PCA
# each row is a respondent, each column is a PC
# wrap in as.data.frame() so we can use dplyr/tidyr on it

# i.e. focus on row 1
# [1,1] = respondent 1's actual values dot.product pc1 - this is the project of r1 onto pc1 axis
# [1,2] = respodent 1's actual values dot.product pc2 - this is the projection of r2 onto pc2 axis
pc_scores_imp <- as.data.frame(pca_imp$x)

# add a respondent ID column — just row numbers 1 to 905
# useful if you later want to join these scores back to the original data
pc_scores_imp$id <- 1:nrow(pc_scores_imp)

# We expect each respondent to have 17 scores + 1 id column
dim(pc_scores_imp) == c(905, 18)

# turn into long format
library(tidyr)

# pivot_longer reshapes from wide to long:
# BEFORE: 905 rows x 18 columns (one column per PC)
# AFTER:  905 x 17 = 15,385 rows, with two key columns:
#           "PC"    — which PC this row refers to (e.g. "PC1", "PC2")
#           "score" — that respondent's score on that PC
# this is necessary because ggplot expects one row per observation to facet by PC
pc_long <- pc_scores_imp %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "PC",
    values_to = "score"
  )

# histogram
ggplot(pc_long %>% filter(PC %in% paste0("PC", 1:9)),
       aes(x = score)) +
  geom_histogram(bins = 30) +
  facet_wrap(~PC, scales = "free") +
  theme_minimal() +
  labs(
    title = "Distribution of First 9 Principal Components (Imputed Data)",
    x = "PC score",
    y = "Count"
  )

ggplot(pc_long %>% filter(PC %in% paste0("PC", 10:17)),
       aes(x = score)) +
  geom_histogram(bins = 30) +
  facet_wrap(~PC, scales = "free") +
  theme_minimal() +
  labs(
    title = "Distribution of 10-19 Principal Components (Imputed Data)",
    x = "PC score",
    y = "Count"
  )
```


```{r, fig.width=8, fig.height=4.5}
# estimated PC scores from each respondent
pc_scores_ind <- as.data.frame(pca_imp$x[, 1:7])

# connect with wave
pc_scores_ind$Wave <- dat_bin$WAVE_NUM

# mean PC score + wave size by wave
wave_pc_mean <- pc_scores_ind %>%
  filter(!is.na(Wave)) %>%
  group_by(Wave) %>%
  summarise(
    wave_size = n(),
    across(starts_with("PC"), mean, na.rm = TRUE),
    .groups = "drop"
  )

# explained variance by individual PCA
var_exp <- summary(pca_imp)$importance[2, ] * 100

plot_df <- wave_pc_mean %>%
  pivot_longer(
    cols = c(PC2, PC3, PC4, PC5, PC6, PC7),
    names_to = "PC_y",
    values_to = "y_value"
  ) %>%
  mutate(
    x_value = PC1,
    Wave_num = as.numeric(Wave),
    x_label = paste0("PC1 (", round(var_exp[1], 1), "%)"),
    y_label = case_when(
      PC_y == "PC2" ~ paste0("PC2 (", round(var_exp[2], 1), "%)"),
      PC_y == "PC3" ~ paste0("PC3 (", round(var_exp[3], 1), "%)"),
      PC_y == "PC4" ~ paste0("PC4 (", round(var_exp[4], 1), "%)"),
      PC_y == "PC5" ~ paste0("PC5 (", round(var_exp[5], 1), "%)"),
      PC_y == "PC6" ~ paste0("PC6 (", round(var_exp[6], 1), "%)"),
      PC_y == "PC7" ~ paste0("PC7 (", round(var_exp[7], 1), "%)")
    ),
    panel = paste0(x_label, " vs ", y_label)
  )

# weighted average point for each subplot
avg_df <- plot_df %>%
  group_by(panel) %>%
  summarise(
    x_avg = weighted.mean(x_value, wave_size, na.rm = TRUE),
    y_avg = weighted.mean(y_value, wave_size, na.rm = TRUE),
    .groups = "drop"
  )

wave_breaks <- c(0, 5, 10, 15)

p1 <- ggplot(plot_df, aes(
  x = x_value,
  y = y_value,
  color = Wave_num,
  alpha = Wave_num,
  size = wave_size
)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +

  geom_point() +

  # average black dot
  geom_point(
    data = avg_df,
    aes(x = x_avg, y = y_avg),
    inherit.aes = FALSE,
    color = "black",
    size = 4
  ) +

  geom_text(
    data = plot_df %>% distinct(panel, x_label),
    aes(x = 0, y = -1.8, label = x_label),
    inherit.aes = FALSE,
    size = 3.5
  ) +

  geom_text(
    data = plot_df %>% distinct(panel, y_label),
    aes(x = -1.8, y = 0, label = y_label),
    inherit.aes = FALSE,
    angle = 90,
    size = 3.5
  ) +

  facet_wrap(~panel, ncol = 3) +

  scale_color_gradient(
    low = "#8B0000",
    high = "#FFC0CB",
    breaks = wave_breaks,
    labels = wave_breaks,
    name = "Wave",
    guide = guide_colorbar(reverse = TRUE)
  ) +

  scale_alpha(range = c(1, 0.4), guide = "none") +

  scale_size_continuous(
    name = "Wave size",
    range = c(2.5, 6)
  ) +

  coord_cartesian(
    xlim = c(-2, 2),
    ylim = c(-2, 2)
  ) +

  labs(
    title = "Wave-level Mean PCA Scores (Darker = earlier waves)"
  ) +

  theme_minimal() +
  theme(
    axis.title = element_blank(),
    strip.text = element_blank()
  )
p1
ggsave("path2017_spca.png", plot = p1, width = 8, height = 6, dpi = 300)
```

## Logistic PCA

- Logistic PCA respects binary data by modeling it through a logistic (Bernoulli) likelihood instead of assuming normality.
```{r}
## Logistic PCA
library(logisticPCA)

# Prepare the data — same variables as before but two extra steps:
dat_lpca_imp <- dat_imp_seq %>%
  dplyr::select(all_of(pca_vars)) %>%
  # logisticPCA requires strictly numeric input, so force all columns to numeric
  # (they may currently be integer or factor)
  mutate(across(everything(), as.numeric))

# logisticPCA requires a matrix not a data frame
X_lpca <- as.matrix(dat_lpca_imp)
```

### Choose an optimal value of `m`
```{r}
cv_fit <- cv.lpca(X_lpca, ks =length(pca_vars), ms = 1:10)
plot(cv_fit)
```
### 508
```{r}
cv_fit <- cv.lpca(X_lpca, ks = length(pca_vars), ms = 1:10)

# one-row frame so the reference line draws once (and gets one legend key)
opt_m <- data.frame(xintercept = 3)

p <- plot(cv_fit) +

  # --- NEW: dashed vertical line at m = 3, shown in legend as "Optimal m" ---
  geom_vline(
    data = opt_m,
    aes(xintercept = xintercept, linetype = "Optimal m"),
    colour = "black", linewidth = 0.7, inherit.aes = FALSE
  ) +
  scale_linetype_manual(name = NULL, values = c("Optimal m" = "dashed")) +

  # --- axis + legend titles ---
  labs(
    x      = "Tuning parameter m",
    y      = "Log-likelihood",
    colour = "Number of PCs"
  ) +

  # --- 508: recolour the data line to a high-contrast, colorblind-safe blue ---
  scale_colour_manual(
    values = "#0072B2",
    labels = paste0(length(pca_vars), " PCs")
  ) +

  theme_bw() +
  theme(
    axis.title   = element_text(size = 16),
    axis.text    = element_text(size = 13, colour = "black"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 13)
  )
p
ggsave("path2017_cv_FINAL.png", plot = p, width = 8, height = 5, dpi = 300)

```

```{r}
lpca_imp <- logisticPCA(
  X_lpca,
  k = length(pca_vars), # number of PCs to extract, set to 17 (all variables)
  m = 3 
)

# Select only the score columns (those starting with "V") from the imputed LPCA scores data frame
lpca_scores_imp <- as.data.frame(lpca_imp$PCs)

lpca_scores_only <- lpca_scores_imp %>%
  dplyr::select(starts_with("V"))

# Compute the standard deviation of each score column
lpca_sd <- apply(lpca_scores_only, 2, sd)
# Square the standard deviations to get the variance of each component
lpca_var <- lpca_sd^2
# Compute each component's proportion of total variance
lpca_prop_var <- lpca_var / sum(lpca_var)

# reorder everything by descending variance before computing cumulative sum
var_order <- order(lpca_prop_var, decreasing = TRUE)
lpca_sd       <- lpca_sd[var_order]
lpca_var      <- lpca_var[var_order]
lpca_prop_var <- lpca_prop_var[var_order]

# cumulative sum is now computed on the sorted proportions
lpca_cum_var <- cumsum(lpca_prop_var)

# reorder the score columns to match
lpca_scores_only <- lpca_scores_only[, var_order]
lpca_scores_imp  <- lpca_scores_imp[, var_order]

lpca_summary <- rbind(
  `Standard deviation`    = lpca_sd,
  `Proportion of Variance` = lpca_prop_var,
  `Cumulative Proportion`  = lpca_cum_var
)
round(lpca_summary, 5)
```
```{r}
as.data.frame(lpca_imp$U) %>%
  `rownames<-`(pca_vars)  
```

```{r}
# TODO - relable in SORTED ORDER
## Logistic PCA scores + histogram + plot by wave

## PC scores - rename to match convention used for regular PCA
colnames(lpca_scores_imp) <- gsub("^V", "PC", colnames(lpca_scores_imp))

# Add respondent ID
lpca_scores_imp$id <- 1:nrow(lpca_scores_imp)

## Histogram
# reshape from wide to long - same logic as before with regular PCA scores
# BEFORE: 905 rows x k columns (one column per PC)
# AFTER:  905 x k rows, with "PC" identifying which component and "score" the value
lpca_long <- lpca_scores_imp %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "PC",
    values_to = "score"
  )
# plot histograms for the first 7 PCs only
# same diagnostic purpose as before — check that score distributions
# look reasonable with no extreme outliers or unexpected multimodality
ggplot(lpca_long %>% filter(PC %in% paste0("PC", 1:7)),
       aes(x = score)) +
  geom_histogram(bins = 30) +
  facet_wrap(~PC, scales = "free") +
  theme_minimal() +
  labs(
    title = "Distribution of First 7 Logistic Principal Components",
    x = "Logistic PC score",
    y = "Count"
  )
```

```{r}
# prcomp(scale. = TRUE) ensures that regular PCA scores are standardized - no equivalent option for logisticPCA
# Diagnostic step — check the raw score distributions across PCs
# list(min, max, mean, sd) computes all four statistics for every PC column
# if logistic PCA scores are on wildly different scales you'll see it here
# e.g. PC1 ranging from -10 to 10 while PC3 ranges from -0.5 to 0.5
lpca_scores_imp %>%
  summarise(across(starts_with("PC"), list(
    min = min,
    max = max,
    mean = mean,
    sd = sd
  )))

# Diagnostic step — check the raw score distributions across PCs
# list(min, max, mean, sd) computes all four statistics for every PC column
# if logistic PCA scores are on wildly different scales you'll see it here
# e.g. PC1 ranging from -10 to 10 while PC3 ranges from -0.5 to 0.5
lpca_scores_scaled <- lpca_scores_imp %>%
  mutate(across(starts_with("PC"), ~ as.numeric(scale(.x))))

# head(lpca_scores_scaled)
lpca_scores_scaled %>%
  summarise(across(starts_with("PC"), list(
    min = min,
    max = max,
    mean = mean,
    sd = sd
  )))
```

```{r}
for (v in paste0("V", 1:7)) {
  result <- as.data.frame(lpca_imp$U) %>%
    `rownames<-`(pca_vars) %>%
    filter(abs(.data[[v]]) > 0.3) %>%
    select(all_of(v)) %>%
    mutate(V = round(.data[[v]], 2))
  
  print("-----------------------------------")
  print(paste("Component:", v))
  print(result)
}
```

## Logistic PCA plot (accounting for size of wave)
```{r}
# Use scaled logistic PC scores for visualization
# Select only the first 7 PCs (by variance order established earlier)
pc_scores_ind_lpca <- lpca_scores_scaled %>%
  dplyr::select(1:7)

# Tag each respondent with their wave
pc_scores_ind_lpca$Wave <- dat_bin$WAVE_NUM

# Collapse to wave-level means — one row per wave, one column per PC
wave_lpca_mean <- pc_scores_ind_lpca %>%
  group_by(Wave) %>%
  summarise(
    wave_size = n(),
    across(starts_with("PC"), mean, na.rm = TRUE),
    .groups = "drop"
  )

# Recompute variance explained from unscaled scores using the reordered columns
# (scaling changes the SDs to 1, so we need the original scores for this)
lpca_scores_only <- lpca_scores_imp %>%
  dplyr::select(starts_with("PC"))
lpca_sd <- apply(lpca_scores_only, 2, sd)
lpca_var_exp <- (lpca_sd^2 / sum(lpca_sd^2)) * 100

# PC2 has the highest variance explained so it goes on the x-axis
# Plot PC2 vs each of PC1, PC3, PC4, PC5, PC6, PC7 — one panel each
# pivot_longer creates one row per wave per panel
plot_df_lpca <- wave_lpca_mean %>%
  pivot_longer(
    cols = c(PC1, PC3, PC4:PC7),      # these are the y-axis PCs; PC2 is the x-axis
    names_to = "PC_y",
    values_to = "y_value"
  ) %>%
  mutate(
    x_value = PC2,               # x-axis is always PC2 (highest variance component)
    Wave_num = as.numeric(Wave),
    panel = PC_y,                # panel identifier for facet_wrap
    # x-axis label: PC2's variance explained (same across all panels)
    x_label = paste0("PC2 (", round(lpca_var_exp["PC2"], 1), "%)"),
    # y-axis label varies by panel
    y_label = case_when(
      PC_y == "PC1" ~ paste0("PC1 (", round(lpca_var_exp["PC1"], 1), "%)"),
      PC_y == "PC3" ~ paste0("PC3 (", round(lpca_var_exp["PC3"], 1), "%)"),
      PC_y == "PC4" ~ paste0("PC4 (", round(lpca_var_exp["PC4"], 1), "%)"),
      PC_y == "PC5" ~ paste0("PC5 (", round(lpca_var_exp["PC5"], 1), "%)"),
      PC_y == "PC6" ~ paste0("PC6 (", round(lpca_var_exp["PC6"], 1), "%)"),
      PC_y == "PC7" ~ paste0("PC7 (", round(lpca_var_exp["PC7"], 1), "%)")
    ),
    panel = paste0(x_label, " vs ", y_label)
  )

plot_df_lpca <- plot_df_lpca %>%
  mutate(PC_y = factor(PC_y, levels = c("PC1", "PC3", "PC4", "PC5", "PC6", "PC7")),
         panel = factor(panel, levels = unique(panel[order(PC_y)])))

# weighted average point for each subplot
avg_df <- plot_df_lpca %>%
  group_by(panel) %>%
  summarise(
    x_avg = weighted.mean(x_value, wave_size, na.rm = TRUE),
    y_avg = weighted.mean(y_value, wave_size, na.rm = TRUE),
    .groups = "drop"
  )

wave_breaks <- c(0, 5, 10, 15)

p2 <- ggplot(plot_df_lpca, aes(
  x = x_value,
  y = y_value,
  color = Wave_num,      
  alpha = Wave_num,
  size = wave_size
)) +
  # reference lines at origin to anchor the PC space
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +

  # one dot per wave per panel
  geom_point() +

  # average black dot
  geom_point(
    data = avg_df,
    aes(x = x_avg, y = y_avg),
    inherit.aes = FALSE,
    color = "black",
    size = 4
  ) +
  
  # custom x-axis label (PC2) placed near bottom centre of each panel
  geom_text(
    data = plot_df_lpca %>% distinct(panel, x_label),
    aes(x = 0, y = -1.8, label = x_label),
    inherit.aes = FALSE,
    size = 3.5
  ) +

  # custom y-axis label rotated 90 degrees, placed near left of each panel
  geom_text(
    data = plot_df_lpca %>% distinct(panel, y_label),
    aes(x = -1.8, y = 0, label = y_label),
    inherit.aes = FALSE,
    angle = 90,
    size = 3.5
  ) +

  # one panel per y-axis PC, 3 columns
  facet_wrap(~panel, ncol = 3) +

  # dark red = early waves, pink = later waves
  scale_color_gradient(
    low = "#8B0000",
    high = "#FFC0CB",
    breaks = wave_breaks,
    labels = wave_breaks,
    name = "Wave",
    guide = guide_colorbar(reverse = TRUE)
  ) +

  # alpha fades later waves but needs no legend entry
  scale_alpha(range = c(1, 0.4), guide = "none") +
  
  scale_size_continuous(
    name = "Wave size",
    range = c(2.5, 6)
  ) +

  # fix all panels to same axis limits for comparability
  coord_fixed(
    xlim = c(-2, 2),
    ylim = c(-2, 2)
  ) +

  labs(
    title = "Wave-level Mean Logistic PCA Scores (Darker = earlier waves)"
  ) +

  theme_minimal() +
  theme(
    axis.title = element_blank(),   # suppressed — geom_text labels serve this role
    strip.text = element_blank()    # panel headers removed — y_label identifies each panel
  )
p2
ggsave("path2017_lpca.png", plot = p2)
```

```{r}
sort(pca_vars)
```

```{r}
write.csv(lpca_scores_scaled, "lpca_scores_scaled_PATH2017.csv")
write.csv(lpca_scores_imp, "lpca_scores_imp_PATH2017.csv")
write.csv(dat_bin, "dat_bin_PATH2017.csv")
```