---
title: "PCA (HAWK)-05.08.2026"
author: "Angelina"
date: "2026-02-16"
output: html_document
---

## Data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, include=FALSE}
if (!require("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(
  dplyr, tidyr, ggplot2, ggforce, stringr, forcats, purrr,
  janitor, skimr, gtsummary, writexl, vcd, reshape2
)
```

```{r}
HAWK <- read.csv("O:/Korean Health/Data/Codebook/Recoding/Data/DATA_ALL (5.20.2025).csv") %>%
  filter(!(ELIGIBLE_COMPLETE!=1))
dim(HAWK)
df <- HAWK
```
```{r}
table(df$G5, useNA = "ifany")
```

## Variable choice
```{r}
var_list <- c(
  # grouping
  "WAVE", "SEED",

  # Sociodemographic
  "AGE",
  "B1_MALE",
  "B4", # education level
  "N1", # employment status
  "B2", # marital status
  "N3", # income

  # Immigration
  "HOMELANG",
  "SC5_EN",
  "SC5_KR",
  "G5", # citizenship/green card related
  "G2", # year move to US
  "G3", # ethnic identity
  "G1", # country of birth
  "H6", # politics interest

  # Health
  "BMI_CAT", # BMI category
  "ANY_DIFFICULTY", # any difficulty (yes/no)
  "C1_0", "C1_1",
  "E5_1_1", "E5_1_2", "E5_EX_1","E5_EX_2", # life satisfaction item
  "K6_CAT", # psychological distress category
  "LONELINESS_CAT", # loneliness yes/no
  "INSCOV", "M2", # times seeing a medical doctor (continuous)

  # Social network
  "P1_KOREAN", # number of Korean adults known
  "NUM_CLOSE_KR",  # number of Korean people feel close to
  "NUM_CLOSE",
  "H4_ANSWERED" # org membership answered indicator
)

var_list_exist <- var_list[var_list %in% names(df)]
missing_cols <- setdiff(var_list, var_list_exist)
if(length(missing_cols) > 0){
  message("These variables are not found in df: ", paste(missing_cols, collapse = ", "))
}

survey_year <- 2023

dat <- df %>%
  dplyr::select(all_of(var_list_exist)) %>%
  dplyr::mutate(
    # cap P1_KOREAN at 100
    P1_KOREAN = if ("P1_KOREAN" %in% names(.)) pmin(P1_KOREAN, 100) else P1_KOREAN,

    # years in US (structural skip for G2)
    years_in_us = case_when(
      "G1" %in% names(.) & G1 == 1 ~ AGE,
      "G1" %in% names(.) & "G2" %in% names(.) & G1 != 1 & !is.na(G2) ~ survey_year - G2,
      TRUE ~ NA_real_
    ),

    # doctor visits eligibility (structural skip via INSCOV)
    doctor_visits = case_when(
      "INSCOV" %in% names(.) & INSCOV == 1 ~ M2,
      "INSCOV" %in% names(.) & INSCOV == 0 ~ 0,
      TRUE ~ NA_real_
    ),
    doctor_eligible = case_when(
      "INSCOV" %in% names(.) & INSCOV == 1 ~ 1,
      "INSCOV" %in% names(.) & INSCOV == 0 ~ 0,
      TRUE ~ NA_real_
    )
  )

# (A) Self-rated health combine
if (all(c("C1_0","C1_1") %in% names(dat))) {
  dat <- dat %>%
    mutate(
      C1_0_4 = case_when(
        C1_0 %in% c(1,2) ~ 1,
        C1_0 == 3 ~ 2,
        C1_0 == 4 ~ 3,
        C1_0 == 5 ~ 4,
        TRUE ~ NA_real_
      ),
      C1_1_4 = case_when(
        C1_1 %in% 1:4 ~ as.numeric(C1_1),
        TRUE ~ NA_real_
      ),
      self_rated_health_4 = coalesce(C1_0_4, C1_1_4),
      self_rated_health_4_f = factor(
        self_rated_health_4,
        levels = 1:4,
        labels = c("Excellent/Very good", "Good", "Fair / Neither", "Poor / Bad")
      )
    )
}

# (B) Life satisfaction composite
have_e5 <- all(c("E5_1_1","E5_1_2","E5_EX_1","E5_EX_2") %in% names(dat))
if (have_e5) {
  dat <- dat %>%
    mutate(
      E5_1_1_01  = (E5_1_1  - 1) / 4,
      E5_EX_1_01 = (E5_EX_1 - 1) / 4,
      E5_1_2_01  = (E5_1_2  - 1) / 5,
      E5_EX_2_01 = (E5_EX_2 - 1) / 5
    ) %>%
    rowwise() %>%
    mutate(
      life_satisfaction_01 = mean(
        c_across(c(E5_1_1_01, E5_1_2_01, E5_EX_1_01, E5_EX_2_01)),
        na.rm = TRUE
      )
    ) %>%
    ungroup() %>%
    mutate(
      life_satisfaction_01 = if_else(is.nan(life_satisfaction_01), NA_real_, life_satisfaction_01),
      life_satisfaction_100 = 100 * life_satisfaction_01
    )
}

# (C) Citizenship combine
if (all(c("G1", "G5") %in% names(dat))) {
  dat <- dat %>%
    mutate(
      us_citizen_combined = case_when(
        G1 == 1 ~ 1,
        G1 != 1 & G5 == 1 ~ 1,
        G1 != 1 & G5 %in% c(2, 3) ~ 0,
        TRUE ~ NA_real_
      ),
      us_citizen_combined_f = factor(
        us_citizen_combined,
        levels = c(0, 1),
        labels = c("Not a US citizen", "US citizen")
      )
    )
}

# (D) Ethnic identity gate: G3 asked only if G1 != 1 (non-US born)
if (all(c("G1","G3") %in% names(dat))) {

  dat <- dat %>%
    mutate(
      g3_eligible = case_when(
        G1 == 1 ~ 0, # US-born -> not asked G3 by design
        G1 != 1 ~ 1, # non-US born -> eligible
        TRUE ~ NA_real_
      ),

      korean_identity = case_when(
        G1 == 1 ~ 0,
        G1 != 1 & G3 == 2 ~ 1, # Korean
        G1 != 1 & G3 %in% c(1,3,4) ~ 0, # non-Korean categories
        TRUE ~ NA_real_
      ),

      korean_identity_f = factor(
        korean_identity,
        levels = c(0,1),
        labels = c("Non-Korean identity","Korean")
      )
    )
}

# drop raw items you no longer want to show
dat <- dat %>%
  select(
    -any_of(c(
      "C1_0","C1_1","C1_0_4","C1_1_4",
      "E5_1_1","E5_1_2","E5_EX_1","E5_EX_2",
      "E5_1_1_01","E5_1_2_01","E5_EX_1_01","E5_EX_2_01",
      "life_satisfaction_01",
      "G5","G2","INSCOV","M2","G3"
    ))
  )

## cut points
income_cut <- 30000
years_us_cut <- 13 # survey year (2023) - 2010
life_sat_med <- median(dat$life_satisfaction_100, na.rm = TRUE)

dat_bin <- dat %>%
  dplyr::mutate(
    seed01 = dplyr::case_when(
      SEED == 1 ~ 1,
      SEED == 0 ~ 0,
      TRUE ~ NA_real_
    ),
    seed_group = factor(seed01, levels = c(0, 1), labels = c("Non-seed", "Seed")),

    ## sociodemographic
    age_ge35 = dplyr::case_when(
      AGE >= 35 ~ 1,
      !is.na(AGE) ~ 0,
      TRUE ~ NA_real_
    ),

    male = dplyr::case_when(
      B1_MALE == 0 ~ 1,
      B1_MALE == 1 ~ 0,
      TRUE ~ NA_real_
    ),

    somecollege_plus = dplyr::case_when(
      B4 >= 4 ~ 1,
      !is.na(B4) ~ 0,
      TRUE ~ NA_real_
    ),

    employed = dplyr::case_when(
      N1 %in% c(1, 2) ~ 1,
      !is.na(N1) ~ 0,
      TRUE ~ NA_real_
    ),

    married_partnered = dplyr::case_when(
      B2 %in% c(1, 2) ~ 1,
      !is.na(B2) ~ 0,
      TRUE ~ NA_real_
    ),

    income_ge30000 = dplyr::case_when(
      N3 >= income_cut ~ 1,
      !is.na(N3) ~ 0,
      TRUE ~ NA_real_
    ),

    ## immigration / language
    home_lang_korean = dplyr::case_when(
      HOMELANG == 2 ~ 1,
      !is.na(HOMELANG) ~ 0,
      TRUE ~ NA_real_
    ),

    english_proficient = dplyr::case_when(
      SC5_EN >= 4 ~ 1,
      !is.na(SC5_EN) ~ 0,
      TRUE ~ NA_real_
    ),

    korean_proficient = dplyr::case_when(
      SC5_KR >= 4 ~ 1,
      !is.na(SC5_KR) ~ 0,
      TRUE ~ NA_real_
    ),

    us_citizen = dplyr::case_when(
      us_citizen_combined == 1 ~ 1,
      us_citizen_combined == 0 ~ 0,
      TRUE ~ NA_real_
    ),

    years_us_recent = dplyr::case_when(
      years_in_us >= years_us_cut ~ 1,
      !is.na(years_in_us) ~ 0,
      TRUE ~ NA_real_
    ),

    korean_identity = dplyr::case_when(
      korean_identity == 1 ~ 1,
      korean_identity == 0 ~ 0,
      TRUE ~ NA_real_
    ),

    us_born = dplyr::case_when(
      G1 == 1 ~ 1,
      !is.na(G1) ~ 0,
      TRUE ~ NA_real_
    ),

    interestin_Korean_politics = dplyr::case_when(
      H6 %in% c(2, 3) ~ 1,
      H6 %in% c(1, 4) ~ 0,
      TRUE ~ NA_real_
    ),

    ## health
    overweight_obese = dplyr::case_when(
      BMI_CAT %in% c(3, 4) ~ 1,
      BMI_CAT %in% c(1, 2) ~ 0,
      TRUE ~ NA_real_
    ),

    any_difficulty = dplyr::case_when(
      ANY_DIFFICULTY == 1 ~ 1,
      ANY_DIFFICULTY == 0 ~ 0,
      TRUE ~ NA_real_
    ),

    srh_fairpoor = dplyr::case_when(
      self_rated_health_4 %in% c(3, 4) ~ 1,
      self_rated_health_4 %in% c(1, 2) ~ 0,
      TRUE ~ NA_real_
    ),

    life_satisfaction_low = dplyr::case_when(
      life_satisfaction_100 < life_sat_med ~ 1,
      !is.na(life_satisfaction_100) ~ 0,
      TRUE ~ NA_real_
    ),

    distress_high = dplyr::case_when(
      K6_CAT >= 2 ~ 1,
      !is.na(K6_CAT) ~ 0,
      TRUE ~ NA_real_
    ),

    lonely = dplyr::case_when(
      LONELINESS_CAT == 1 ~ 1,
      LONELINESS_CAT == 0 ~ 0,
      TRUE ~ NA_real_
    ),

    doctor_visit_any = dplyr::case_when(
      doctor_visits > 0 ~ 1,
      !is.na(doctor_visits) ~ 0,
      TRUE ~ NA_real_
    ),

    ## social network
    korean_known_gt10 = dplyr::case_when(
      P1_KOREAN > 10 ~ 1,
      !is.na(P1_KOREAN) ~ 0,
      TRUE ~ NA_real_
    ),

    close_korean_any = dplyr::case_when(
      NUM_CLOSE_KR > 0 ~ 1,
      !is.na(NUM_CLOSE_KR) ~ 0,
      TRUE ~ NA_real_
    ),

    close_people_any = dplyr::case_when(
      NUM_CLOSE > 0 ~ 1,
      !is.na(NUM_CLOSE) ~ 0,
      TRUE ~ NA_real_
    ),

    org_membership_answered = dplyr::case_when(
      H4_ANSWERED == 1 ~ 1,
      H4_ANSWERED == 0 ~ 0,
      TRUE ~ NA_real_
    )
  )
```

## correlation test
```{r}
# Sociodemographic
cor(dat_bin[, c("somecollege_plus", "income_ge30000", "employed")], use = "pairwise.complete.obs")

# immigration
cor(dat_bin[, c("home_lang_korean", "english_proficient", "korean_proficient", "us_citizen", "years_us_recent", "korean_identity", "interestin_Korean_politics")], use = "pairwise.complete.obs")

# Physical Health
cor(dat_bin[, c("overweight_obese", "srh_fairpoor")], use = "pairwise.complete.obs")

# Mental Health
cor(dat_bin[, c("any_difficulty", "life_satisfaction_low", "distress_high", "lonely")], use = "pairwise.complete.obs")

# Social Network
cor(dat_bin[, c("korean_known_gt10", "close_korean_any", "close_people_any", "org_membership_answered")], use = "pairwise.complete.obs")
dat_bin %>% count(org_membership_answered)
cor(dat_bin[, c("korean_known_gt10", "close_korean_any", "close_people_any")], use = "pairwise.complete.obs")
```

## Missingness
```{r}
dat_miss_bin <- dat_bin %>%
  dplyr::select(
    seed01, seed_group,
    
    # Sociodemograpgic
    age_ge35,
    male,
    employed,
    married_partnered,
    income_ge30000,

    # Immigrant
    english_proficient,
    years_us_recent,
    korean_identity,
    interestin_Korean_politics,
    
    # Physical Health
    overweight_obese,
    any_difficulty,
    srh_fairpoor,
    doctor_visit_any,
    
    # Mental Health
    life_satisfaction_low,
    distress_high,
    lonely,

    # Social Network
    korean_known_gt10,
    close_korean_any,
  )

var_labels_bin <- c(
  # Sociodemographic
  age_ge35 = "age >= 35",
  male = "male",
  employed = "currently employed",
  married_partnered = "married or living with partner",
  income_ge30000 = "income >= 30000",

  # Immigrant
  english_proficient = "English proficiency: well or very well",
  years_us_recent = "years in US >= 13",
  korean_identity = "Korean identity",
  us_born = "born in US",
  interestin_Korean_politics = "more or equal interest in Korean politics",

  # Health
  overweight_obese = "BMI >= 25 (overweight or obese)",
  any_difficulty = "any difficulty in functioning",
  srh_fairpoor = "self-rated health fair/poor",
  life_satisfaction_low = "below median life satisfaction",
  distress_high = "moderate or serious psychological distress",
  lonely = "loneliness = yes",
  doctor_visit_any = "visited doctor at least once",

  # Social network
  korean_known_gt10 = "know > 10 Korean adults",
  close_korean_any = "close Korean contacts > 0"
)
```

## Plot
```{r}
analysis_vars_bin <- setdiff(names(dat_miss_bin), c("seed01", "seed_group"))

missing_overall_bin <- dat_miss_bin %>%
  dplyr::summarise(across(all_of(analysis_vars_bin), ~mean(is.na(.)) * 100)) %>%
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") %>%
  dplyr::mutate(variable_label = dplyr::recode(variable, !!!var_labels_bin)) %>%
  dplyr::arrange(desc(pct_missing))

missing_overall_bin

ggplot(missing_overall_bin,
       aes(x = reorder(variable_label, pct_missing), y = pct_missing)) +
  geom_col() +
  coord_flip() +
  labs(
    x = NULL,
    y = "% Missing",
    title = "Missingness by binary variables (overall)"
  ) +
  theme_minimal()
```

## Table (for imputation understanding)
```{r}
missing_table_bin <- missing_overall_bin %>%
  dplyr::select(variable, variable_label, pct_missing) %>%
  dplyr::mutate(
    pct_missing = round(pct_missing, 2)
  ) %>%
  dplyr::arrange(desc(pct_missing))

missing_table_bin
```

# PCA with imputation
## PCA variable
```{r}
## seed01 could be predictor due to it don't have missing value, but not putting in PCA main analysis
pca_vars <- setdiff(names(dat_miss_bin), c("WAVE", "seed01", "seed_group"))
imp_vars <- setdiff(names(dat_miss_bin), c("WAVE", "seed_group"))

## missing rate
miss_rate <- sapply(dat_miss_bin[imp_vars], function(x) mean(is.na(x)) * 100)

miss_df <- data.frame(
  variable = names(miss_rate),
  pct_missing = as.numeric(miss_rate)
) %>%
  arrange(pct_missing)

miss_df
```

## Hot deck
```{r}
## one-variable hot deck imputation
hotdeck_one_var <- function(data, target, predictors, k = 5, seed = 123) {
  set.seed(seed)

  x <- data

  miss_idx <- which(is.na(x[[target]]))
  obs_idx  <- which(!is.na(x[[target]]))

  if (length(miss_idx) == 0) return(x)

  if (length(predictors) == 0) {
    x[[target]][miss_idx] <- sample(x[[target]][obs_idx],
                                    size = length(miss_idx),
                                    replace = TRUE)
    return(x)
  }

  donor_mat <- as.matrix(x[obs_idx, predictors, drop = FALSE])
  recip_mat <- as.matrix(x[miss_idx, predictors, drop = FALSE])

  ## numeric
  donor_mat <- apply(donor_mat, 2, as.numeric)
  recip_mat <- apply(recip_mat, 2, as.numeric)

  ## If there is only one predictor, put in matrix
  if (is.null(dim(donor_mat))) donor_mat <- matrix(donor_mat, ncol = 1)
  if (is.null(dim(recip_mat))) recip_mat <- matrix(recip_mat, ncol = 1)

  donor_y <- x[[target]][obs_idx]

  imputed_values <- sapply(seq_len(nrow(recip_mat)), function(i) {
    d <- rowSums((t(t(donor_mat) - recip_mat[i, ]))^2)

    ## closest donor k
    k_use <- min(k, length(d))
    nearest_idx <- order(d)[1:k_use]

    sample(donor_y[nearest_idx], size = 1)
  })

  x[[target]][miss_idx] <- imputed_values
  x
}
```

## sequential hot deck from lowest missing rate
```{r}
dat_imp_seq <- dat_miss_bin

## let 0% missing be in the predictor pool
predictor_pool <- miss_df %>%
  filter(pct_missing == 0) %>%
  pull(variable)

## variables for sequentially impute
target_order <- miss_df %>%
  filter(pct_missing > 0) %>%
  arrange(pct_missing) %>%
  pull(variable)

predictor_pool
target_order
```
```{r}
## sequential hot deck
imputation_log <- data.frame(
  step = integer(),
  target = character(),
  pct_missing = numeric(),
  n_predictors_used = integer(),
  predictors_used = character()
)

for (j in seq_along(target_order)) {
  target_j <- target_order[j]

  ## predictors couldn't include themselves
  current_predictors <- setdiff(predictor_pool, target_j)

  dat_imp_seq <- hotdeck_one_var(
    data = dat_imp_seq,
    target = target_j,
    predictors = current_predictors,
    k = 5,
    seed = 100 + j
  )

  ## adding target_j in the predictor pool after imputing
  predictor_pool <- c(predictor_pool, target_j)

  imputation_log <- bind_rows(
    imputation_log,
    data.frame(
      step = j,
      target = target_j,
      pct_missing = miss_df$pct_missing[match(target_j, miss_df$variable)],
      n_predictors_used = length(current_predictors),
      predictors_used = paste(current_predictors, collapse = ", ")
    )
  )
}

imputation_log
```

## Check
```{r}
colSums(is.na(dat_imp_seq))
sum(is.na(dat_imp_seq[pca_vars]))
```
## PCA
```{r}
## imputed data for PCA
dat_pca_imp <- dat_imp_seq %>%
  dplyr::select(all_of(pca_vars))

dim(dat_pca_imp)

pca_imp <- prcomp(dat_pca_imp, center = TRUE, scale. = TRUE)

summary(pca_imp)
```

# PCA with imputation
```{r}
round(pca_imp$rotation,2)
```

```{r}
rot <- round(pca_imp$rotation, 2)[, 1:7]

lapply(colnames(rot), function(pc) {
  rot[abs(rot[, pc]) > 0.3, pc, drop = FALSE]
})
```

## PC distribution (histogram)
```{r}
# PC scores (each row = individual, each column = PC)
pc_scores_imp <- as.data.frame(pca_imp$x)

pc_scores_imp$id <- 1:nrow(pc_scores_imp)

dim(pc_scores_imp)

# turn into long format
library(tidyr)

pc_long <- pc_scores_imp %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "PC",
    values_to = "score"
  )

# histogram
ggplot(pc_long %>% filter(PC %in% paste0("PC", 1:9)),
       aes(x = score)) +
  geom_histogram(bins = 30) +
  facet_wrap(~PC, scales = "free") +
  theme_minimal() +
  labs(
    title = "Distribution of First 9 Principal Components (Imputed Data)",
    x = "PC score",
    y = "Count"
  )

ggplot(pc_long %>% filter(PC %in% paste0("PC", 10:18)),
       aes(x = score)) +
  geom_histogram(bins = 30) +
  facet_wrap(~PC, scales = "free") +
  theme_minimal() +
  labs(
    title = "Distribution of 10-19 Principal Components (Imputed Data)",
    x = "PC score",
    y = "Count"
  )
```

## Plot by wave
from  mean variables by wave -> PCA to PCA on individuals -> mean PC scores by wave
```{r, fig.width=8, fig.height=4.5}
# estimated PC scores from each respondent
pc_scores_ind <- as.data.frame(pca_imp$x[, 1:7])

# connect with wave
pc_scores_ind$Wave <- dat_bin$WAVE

# mean PC score by wave
wave_pc_mean <- pc_scores_ind %>%
  group_by(Wave) %>%
  summarise(
    across(starts_with("PC"), mean, na.rm = TRUE),
    .groups = "drop"
  )

# explained variance by individual PCA
var_exp <- summary(pca_imp)$importance[2, ] * 100

plot_df <- wave_pc_mean %>%
  pivot_longer(
    cols = c(PC2, PC3, PC4, PC5, PC6, PC7),
    names_to = "PC_y",
    values_to = "y_value"
  ) %>%
  mutate(
    x_value = PC1,
    panel = dplyr::case_when(
      PC_y == "PC2" ~ paste0("PC1 (", round(var_exp[1], 1), "%) vs PC2 (", round(var_exp[2], 1), "%)"),
      PC_y == "PC3" ~ paste0("PC1 (", round(var_exp[1], 1), "%) vs PC3 (", round(var_exp[3], 1), "%)"),
      PC_y == "PC4" ~ paste0("PC1 (", round(var_exp[1], 1), "%) vs PC4 (", round(var_exp[4], 1), "%)"),
      PC_y == "PC5" ~ paste0("PC1 (", round(var_exp[1], 1), "%) vs PC5 (", round(var_exp[5], 1), "%)"),
      PC_y == "PC6" ~ paste0("PC1 (", round(var_exp[1], 1), "%) vs PC6 (", round(var_exp[6], 1), "%)"),
      PC_y == "PC7" ~ paste0("PC1 (", round(var_exp[1], 1), "%) vs PC7 (", round(var_exp[7], 1), "%)")
    )
  )

# zero in the  middle
x_lim <- max(abs(plot_df$x_value), na.rm = TRUE)
y_lim <- max(abs(plot_df$y_value), na.rm = TRUE)

wave_breaks <- seq(min(plot_df$Wave), max(plot_df$Wave), by = 2)

ggplot(plot_df, aes(
  x = x_value,
  y = y_value,
  color = as.numeric(Wave),
  alpha = as.numeric(Wave)
)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 4) +
  facet_wrap(~panel, ncol = 3) +
  scale_color_gradientn(
    colours = c("#8B0000", "#FFC0CB"),
    breaks = wave_breaks,
    labels = wave_breaks,
    name = "Wave"
  ) +
  scale_alpha(range = c(1, 0.4), guide = "none") +
  coord_cartesian(
    xlim = c(-x_lim, x_lim),
    ylim = c(-y_lim, y_lim)
  ) +
  labs(
    x = paste0("Mean PC1 score (", round(var_exp[1], 1), "%)"),
    y = "Mean PC score",
    title = "Wave-level Mean PCA Scores (Darker = earlier waves)"
  ) +
  theme_minimal()
```
```{r}
plot_df <- plot_df %>%
  mutate(
    x_label = paste0("PC1 (", round(var_exp[1], 1), "%)"),
    y_label = case_when(
      PC_y == "PC2" ~ paste0("PC2 (", round(var_exp[2], 1), "%)"),
      PC_y == "PC3" ~ paste0("PC3 (", round(var_exp[3], 1), "%)"),
      PC_y == "PC4" ~ paste0("PC4 (", round(var_exp[4], 1), "%)"),
      PC_y == "PC5" ~ paste0("PC5 (", round(var_exp[5], 1), "%)"),
      PC_y == "PC6" ~ paste0("PC6 (", round(var_exp[6], 1), "%)"),
      PC_y == "PC7" ~ paste0("PC7 (", round(var_exp[7], 1), "%)")
    )
  )

ggplot(plot_df, aes(
  x = x_value,
  y = y_value,
  color = as.numeric(Wave),
  alpha = as.numeric(Wave)
)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +

  geom_point(size = 4) +
  
  geom_text(
    data = plot_df %>% distinct(panel, x_label),
    aes(x = 0, y = -1.8, label = x_label),
    inherit.aes = FALSE,
    size = 3.5
  ) +

  geom_text(
    data = plot_df %>% distinct(panel, y_label),
    aes(x = -1.8, y = 0, label = y_label),
    inherit.aes = FALSE,
    angle = 90,
    size = 3.5
  ) +

  facet_wrap(~panel, ncol = 3) +

  scale_color_gradientn(
    colours = c("#8B0000", "#FFC0CB"),
    breaks = wave_breaks,
    labels = wave_breaks,
    name = "Wave"
  ) +

  scale_alpha(range = c(1, 0.4), guide = "none") +

  coord_cartesian(
    xlim = c(-2, 2),
    ylim = c(-2, 2)
  ) +

  labs(
    title = "Wave-level Mean PCA Scores (Darker = earlier waves)"
  ) +

  theme_minimal() +
  theme(
    axis.title = element_blank(),
    strip.text = element_blank()
  )
```
```{r, fig.width=8, fig.height=4.5}
# estimated PC scores from each respondent
pc_scores_ind <- as.data.frame(pca_imp$x[, 1:7])

# connect with wave
pc_scores_ind$Wave <- dat_bin$WAVE

# mean PC score + wave size by wave
wave_pc_mean <- pc_scores_ind %>%
  filter(!is.na(Wave)) %>%
  group_by(Wave) %>%
  summarise(
    wave_size = n(),
    across(starts_with("PC"), mean, na.rm = TRUE),
    .groups = "drop"
  )

# explained variance by individual PCA
var_exp <- summary(pca_imp)$importance[2, ] * 100

plot_df <- wave_pc_mean %>%
  pivot_longer(
    cols = c(PC2, PC3, PC4, PC5, PC6, PC7),
    names_to = "PC_y",
    values_to = "y_value"
  ) %>%
  mutate(
    x_value = PC1,
    Wave_num = as.numeric(Wave),
    x_label = paste0("PC1 (", round(var_exp[1], 1), "%)"),
    y_label = case_when(
      PC_y == "PC2" ~ paste0("PC2 (", round(var_exp[2], 1), "%)"),
      PC_y == "PC3" ~ paste0("PC3 (", round(var_exp[3], 1), "%)"),
      PC_y == "PC4" ~ paste0("PC4 (", round(var_exp[4], 1), "%)"),
      PC_y == "PC5" ~ paste0("PC5 (", round(var_exp[5], 1), "%)"),
      PC_y == "PC6" ~ paste0("PC6 (", round(var_exp[6], 1), "%)"),
      PC_y == "PC7" ~ paste0("PC7 (", round(var_exp[7], 1), "%)")
    ),
    panel = paste0(x_label, " vs ", y_label)
  )

# weighted average point for each subplot
avg_df <- plot_df %>%
  group_by(panel) %>%
  summarise(
    x_avg = weighted.mean(x_value, wave_size, na.rm = TRUE),
    y_avg = weighted.mean(y_value, wave_size, na.rm = TRUE),
    .groups = "drop"
  )

wave_breaks <- c(0, 5, 10, 15)

p1 <- ggplot(plot_df, aes(
  x = x_value,
  y = y_value,
  color = Wave_num,
  alpha = Wave_num,
  size = wave_size
)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +

  geom_point() +

  # average black dot
  geom_point(
    data = avg_df,
    aes(x = x_avg, y = y_avg),
    inherit.aes = FALSE,
    color = "black",
    size = 4
  ) +

  geom_text(
    data = plot_df %>% distinct(panel, x_label),
    aes(x = 0, y = -1.8, label = x_label),
    inherit.aes = FALSE,
    size = 3.5
  ) +

  geom_text(
    data = plot_df %>% distinct(panel, y_label),
    aes(x = -1.8, y = 0, label = y_label),
    inherit.aes = FALSE,
    angle = 90,
    size = 3.5
  ) +

  facet_wrap(~panel, ncol = 3) +

  scale_color_gradient(
    low = "#8B0000",
    high = "#FFC0CB",
    breaks = wave_breaks,
    labels = wave_breaks,
    name = "Wave",
    guide = guide_colorbar(reverse = TRUE)
  ) +

  scale_alpha(range = c(1, 0.4), guide = "none") +

  scale_size_continuous(
    name = "Wave size",
    range = c(2.5, 6)
  ) +

  coord_cartesian(
    xlim = c(-2, 2),
    ylim = c(-2, 2)
  ) +

  labs(
    title = "Wave-level Mean PCA Scores (Darker = earlier waves)"
  ) +

  theme_minimal() +
  theme(
    axis.title = element_blank(),
    strip.text = element_blank()
  )
p1
ggsave("hawk_spca.png", plot = p1, width = 8, height = 6, dpi = 300)
```

## Logistic PCA
```{r}
library(logisticPCA)

dat_lpca_imp <- dat_imp_seq %>%
  dplyr::select(all_of(pca_vars)) %>%
  mutate(across(everything(), as.numeric))

dim(dat_lpca_imp)

X_lpca <- as.matrix(dat_lpca_imp)
```

### Choose an optimal value of `m`
```{r}
cv_fit <- cv.lpca(X_lpca, ks =length(pca_vars), ms = 1:10)
plot(cv_fit)
```

### 508
```{r}
cv_fit <- cv.lpca(X_lpca, ks = length(pca_vars), ms = 1:10)

# one-row frame so the reference line draws once (and gets one legend key)
opt_m <- data.frame(xintercept = 3)

p <- plot(cv_fit) +

  # --- NEW: dashed vertical line at m = 3, shown in legend as "Optimal m" ---
  geom_vline(
    data = opt_m,
    aes(xintercept = xintercept, linetype = "Optimal m"),
    colour = "black", linewidth = 0.7, inherit.aes = FALSE
  ) +
  scale_linetype_manual(name = NULL, values = c("Optimal m" = "dashed")) +

  # --- axis + legend titles ---
  labs(
    x      = "Tuning parameter m",
    y      = "Log-likelihood",
    colour = "Number of PCs"
  ) +

  # --- 508: recolour the data line to a high-contrast, colorblind-safe blue ---
  scale_colour_manual(
    values = "#0072B2",
    labels = paste0(length(pca_vars), " PCs")
  ) +

  theme_bw() +
  theme(
    axis.title   = element_text(size = 16),
    axis.text    = element_text(size = 13, colour = "black"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 13)
  )
p
ggsave("pathHAWK_cv_FINAL.png", plot = p, width = 8, height = 5, dpi = 300)

```

```{r}
lpca_imp <- logisticPCA(
  X_lpca,
  k = length(pca_vars),
  m = 3
)

lpca_imp

## Logistic PCA summary, similar to summary(pca_imp)
lpca_scores_imp <- as.data.frame(lpca_imp$PCs)

colnames(lpca_scores_imp) <- paste0("PC", 1:ncol(lpca_scores_imp))

lpca_scores_imp$id <- 1:nrow(lpca_scores_imp)

lpca_scores_only <- lpca_scores_imp %>%
  dplyr::select(starts_with("PC"))

lpca_sd <- apply(lpca_scores_only, 2, sd)

lpca_var <- lpca_sd^2
lpca_prop_var <- lpca_var / sum(lpca_var)
lpca_cum_var <- cumsum(lpca_prop_var)

lpca_summary <- rbind(
  `Standard deviation` = lpca_sd,
  `Proportion of Variance` = lpca_prop_var,
  `Cumulative Proportion` = lpca_cum_var
)

round(lpca_summary, 5)

## Logistic PCA loadings
lpca_rotation <- lpca_imp$U
rownames(lpca_rotation) <- pca_vars
colnames(lpca_rotation) <- paste0("PC", 1:ncol(lpca_rotation))

round(lpca_rotation, 2)
```

## Logistic PCA loadings check
```{r}
load_lpca_pc1 <- data.frame(
  variable = rownames(lpca_rotation),
  PC1 = lpca_rotation[, 1]
)

load_lpca_pc1 %>%
  arrange(desc(abs(PC1))) %>%
  head(10)

load_lpca_pc2 <- data.frame(
  variable = rownames(lpca_rotation),
  PC2 = lpca_rotation[, 2]
)

load_lpca_pc2 %>%
  arrange(desc(abs(PC2))) %>%
  head(10)
```

## Logistic PC distribution (histogram)
```{r}
lpca_scores_scaled <- lpca_scores_imp %>%
  mutate(across(starts_with("PC"), ~ as.numeric(scale(.x))))

lpca_long <- lpca_scores_scaled %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "PC",
    values_to = "score"
  )

ggplot(lpca_long %>% filter(PC %in% paste0("PC", 1:9)),
       aes(x = score)) +
  geom_histogram(bins = 30) +
  facet_wrap(~PC, scales = "free") +
  theme_minimal() +
  labs(
    title = "Distribution of First 9 Logistic Principal Components (Imputed Data)",
    x = "Logistic PC score",
    y = "Count"
  )

ggplot(lpca_long %>% filter(PC %in% paste0("PC", 10:length(pca_vars))),
       aes(x = score)) +
  geom_histogram(bins = 30) +
  facet_wrap(~PC, scales = "free") +
  theme_minimal() +
  labs(
    title = "Distribution of Remaining Logistic Principal Components (Imputed Data)",
    x = "Logistic PC score",
    y = "Count"
  )

```

## Logistic PCA plot by wave
```{r, fig.width=8, fig.height=4.5}
pc_scores_ind_lpca <- lpca_scores_scaled %>%
  dplyr::select(PC1:PC7)

pc_scores_ind_lpca$Wave <- dat_bin$WAVE

wave_lpca_mean <- pc_scores_ind_lpca %>%
  filter(!is.na(Wave)) %>%
  group_by(Wave) %>%
  summarise(
    wave_size = n(),
    across(starts_with("PC"), mean, na.rm = TRUE),
    .groups = "drop"
  )

lpca_var_exp <- lpca_prop_var * 100

plot_df_lpca <- wave_lpca_mean %>%
  pivot_longer(
    cols = c(PC2, PC3, PC4, PC5, PC6, PC7),
    names_to = "PC_y",
    values_to = "y_value"
  ) %>%
  mutate(
    x_value = PC1,
    Wave_num = as.numeric(Wave),
    x_label = paste0("PC1 (", round(lpca_var_exp[1], 1), "%)"),
    y_label = case_when(
      PC_y == "PC2" ~ paste0("PC2 (", round(lpca_var_exp[2], 1), "%)"),
      PC_y == "PC3" ~ paste0("PC3 (", round(lpca_var_exp[3], 1), "%)"),
      PC_y == "PC4" ~ paste0("PC4 (", round(lpca_var_exp[4], 1), "%)"),
      PC_y == "PC5" ~ paste0("PC5 (", round(lpca_var_exp[5], 1), "%)"),
      PC_y == "PC6" ~ paste0("PC6 (", round(lpca_var_exp[6], 1), "%)"),
      PC_y == "PC7" ~ paste0("PC7 (", round(lpca_var_exp[7], 1), "%)")
    ),
    panel = paste0(x_label, " vs ", y_label)
  )

avg_df_lpca <- plot_df_lpca %>%
  group_by(panel) %>%
  summarise(
    x_avg = weighted.mean(x_value, wave_size, na.rm = TRUE),
    y_avg = weighted.mean(y_value, wave_size, na.rm = TRUE),
    .groups = "drop"
  )

wave_breaks <- c(0, 5, 10, 15)

ggplot(plot_df_lpca, aes(
  x = x_value,
  y = y_value,
  color = Wave_num,
  alpha = Wave_num,
  size = wave_size
)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +

  geom_point() +

  geom_point(
    data = avg_df_lpca,
    aes(x = x_avg, y = y_avg),
    inherit.aes = FALSE,
    color = "black",
    size = 4
  ) +

  geom_text(
    data = plot_df_lpca %>% distinct(panel, x_label),
    aes(x = 0, y = -1.8, label = x_label),
    inherit.aes = FALSE,
    size = 3.5
  ) +

  geom_text(
    data = plot_df_lpca %>% distinct(panel, y_label),
    aes(x = -1.8, y = 0, label = y_label),
    inherit.aes = FALSE,
    angle = 90,
    size = 3.5
  ) +

  facet_wrap(~panel, ncol = 3) +

  scale_color_gradient(
    low = "#8B0000",
    high = "#FFC0CB",
    breaks = wave_breaks,
    labels = wave_breaks,
    name = "Wave",
    guide = guide_colorbar(reverse = TRUE)
  ) +

  scale_alpha(range = c(1, 0.4), guide = "none") +

  scale_size_continuous(
    name = "Wave size",
    range = c(2.5, 6)
  ) +

  coord_cartesian(
    xlim = c(-2, 2),
    ylim = c(-2, 2)
  ) +

  labs(
    title = "Wave-level Mean Logistic PCA Scores (Darker = earlier waves)"
  ) +

  theme_minimal() +
  theme(
    axis.title = element_blank(),
    strip.text = element_blank()
  )
  
```

## SORTED Logistic PCA plot by wave
```{r}
# Select only the score columns (those starting with "V") from the imputed LPCA scores data frame
lpca_scores_imp <- as.data.frame(lpca_imp$PCs)

lpca_scores_only <- lpca_scores_imp %>%
  dplyr::select(starts_with("V"))

# Compute the standard deviation of each score column
lpca_sd <- apply(lpca_scores_only, 2, sd)
# Square the standard deviations to get the variance of each component
lpca_var <- lpca_sd^2
# Compute each component's proportion of total variance
lpca_prop_var <- lpca_var / sum(lpca_var)

# reorder everything by descending variance before computing cumulative sum
var_order <- order(lpca_prop_var, decreasing = TRUE)
lpca_sd       <- lpca_sd[var_order]
lpca_var      <- lpca_var[var_order]
lpca_prop_var <- lpca_prop_var[var_order]

# cumulative sum is now computed on the sorted proportions
lpca_cum_var <- cumsum(lpca_prop_var)

# reorder the score columns to match
lpca_scores_only <- lpca_scores_only[, var_order]
lpca_scores_imp  <- lpca_scores_imp[, var_order]

lpca_summary <- rbind(
  `Standard deviation`    = lpca_sd,
  `Proportion of Variance` = lpca_prop_var,
  `Cumulative Proportion`  = lpca_cum_var
)
round(lpca_summary, 5)
```

```{r}
as.data.frame(lpca_imp$U) %>%
  `rownames<-`(pca_vars)  
```

```{r}
# as.data.frame(lpca_imp$U) %>%
#   `rownames<-`(pca_vars) %>%
#   filter(abs(V2) > 0.3) %>%
#   select(V2) %>%
#   mutate(V = round(V2,2))
for (v in paste0("V", 1:7)) {
  result <- as.data.frame(lpca_imp$U) %>%
    `rownames<-`(pca_vars) %>%
    filter(abs(.data[[v]]) > 0.3) %>%
    select(all_of(v)) %>%
    mutate(V = round(.data[[v]], 2))
  
  print("-----------------------------------")
  print(paste("Component:", v))
  print(result)
}
```

```{r}
write.csv(lpca_scores_scaled, "lpca_scores_scaled_HAWK.csv")
write.csv(lpca_scores_imp, "lpca_scores_imp_HAWK.csv")
write.csv(dat_bin, "dat_bin_HAWK.csv")
write.csv(lpca_prop_var, "lpca_prop_var_HAWK.csv")
```

```{r, fig.width=8, fig.height=4.5}
pc_scores_ind_lpca <- lpca_scores_scaled %>%
  dplyr::select(PC1:PC8)

pc_scores_ind_lpca$Wave <- dat_bin$WAVE

wave_lpca_mean <- pc_scores_ind_lpca %>%
  filter(!is.na(Wave)) %>%
  group_by(Wave) %>%
  summarise(
    wave_size = n(),
    across(starts_with("PC"), mean, na.rm = TRUE),
    .groups = "drop"
  )

lpca_var_exp <- lpca_prop_var * 100

plot_df_lpca <- wave_lpca_mean %>%
  pivot_longer(
    cols = c(PC3, PC4, PC5, PC6, PC7, PC1),
    names_to = "PC_y",
    values_to = "y_value"
  ) %>%
  mutate(
    x_value = PC2,
    Wave_num = as.numeric(Wave),
    x_label = paste0("PC2 (", round(lpca_var_exp[1], 1), "%)"),
    y_label = case_when(
      PC_y == "PC3" ~ paste0("PC3 (", round(lpca_var_exp[2], 1), "%)"),
      PC_y == "PC4" ~ paste0("PC4 (", round(lpca_var_exp[3], 1), "%)"),
      PC_y == "PC5" ~ paste0("PC5 (", round(lpca_var_exp[4], 1), "%)"),
      PC_y == "PC6" ~ paste0("PC6 (", round(lpca_var_exp[5], 1), "%)"),
      PC_y == "PC7" ~ paste0("PC7 (", round(lpca_var_exp[6], 1), "%)"),
      PC_y == "PC1" ~ paste0("PC1 (", round(lpca_var_exp[7], 1), "%)")
    ),
    panel = paste0(x_label, " vs ", y_label)
  )

plot_df_lpca <- plot_df_lpca %>%
  mutate(PC_y = factor(PC_y, levels = c("PC3", "PC4", "PC5", "PC6", "PC7", "PC1")),
         panel = factor(panel, levels = unique(panel[order(PC_y)])))

avg_df_lpca <- plot_df_lpca %>%
  group_by(panel) %>%
  summarise(
    x_avg = weighted.mean(x_value, wave_size, na.rm = TRUE),
    y_avg = weighted.mean(y_value, wave_size, na.rm = TRUE),
    .groups = "drop"
  )

wave_breaks <- c(0, 5, 10, 15)

p2 <- ggplot(plot_df_lpca, aes(
  x = x_value,
  y = y_value,
  color = Wave_num,
  alpha = Wave_num,
  size = wave_size
)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +

  geom_point() +

  geom_point(
    data = avg_df_lpca,
    aes(x = x_avg, y = y_avg),
    inherit.aes = FALSE,
    color = "black",
    size = 4
  ) +

  geom_text(
    data = plot_df_lpca %>% distinct(panel, x_label),
    aes(x = 0, y = -1.8, label = x_label),
    inherit.aes = FALSE,
    size = 3.5
  ) +

  geom_text(
    data = plot_df_lpca %>% distinct(panel, y_label),
    aes(x = -1.8, y = 0, label = y_label),
    inherit.aes = FALSE,
    angle = 90,
    size = 3.5
  ) +

  facet_wrap(~panel, ncol = 3) +

  scale_color_gradient(
    low = "#8B0000",
    high = "#FFC0CB",
    breaks = wave_breaks,
    labels = wave_breaks,
    name = "Wave",
    guide = guide_colorbar(reverse = TRUE)
  ) +

  scale_alpha(range = c(1, 0.4), guide = "none") +

  scale_size_continuous(
    name = "Wave size",
    range = c(2.5, 6)
  ) +

  coord_cartesian(
    xlim = c(-2, 2),
    ylim = c(-2, 2)
  ) +

  labs(
    title = "Wave-level Mean Logistic PCA Scores (Darker = earlier waves)"
  ) +

  theme_minimal() +
  theme(
    axis.title = element_blank(),
    strip.text = element_blank()
  )
p2
ggsave("hawk_lpca.png", plot = p2, width = 8, height = 6, dpi = 300)
  
```


```{r}
# write.csv(lpca_scores_scaled, "lpca_scores_scaled_HAWK.csv")
# write.csv(lpca_scores_imp, "lpca_scores_imp_HAWK.csv")
# write.csv(dat_bin, "dat_bin_HAWK.csv")
# write.csv(lpca_prop_var, "lpca_prop_var_HAWK.csv")
```